---
title: "Bjork"
output: html_document
date: "2025-05-11"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Download data

```
kingfisher get --run-identifiers-list seq_accessions.txt --download-threads 2 -m ena-ascp ena-ftp aws-http prefetch

kingfisher get --download-threads 10 -m ena-ascp ena-ftp aws-http prefetch --bioprojects PRJEB70966 PRJEB43119
```

for f in *_1.fastq.gz; do r="${f/_1.fastq.gz/_2.fastq.gz}"; [[ ! -f "$r" ]] && echo "Missing reverse read for: $f"; done


for f in *_2.fastq.gz; do
    r="${f/_2.fastq.gz/_1.fastq.gz}"
    if [[ ! -f "$r" ]]; then
        prefix="${f%_2.fastq.gz}"
        echo "Missing reverse read for: $prefix"
        #echo "kingfisher get -r ${prefix} --force -m ena-ascp ena-ftp aws-http prefetch" >> download_again_cmds.txt
    fi
done

parallel --jobs 10 --halt soon,fail=1 --joblog parallel_redo2_joblog_king.txt < download_again_cmds.txt 2> parallel_redo2_errors_king.log


for f in seq_accessions_part*
do
echo "kingfisher get --run-identifiers-list ${f} --download-threads 2 -m ena-ascp ena-ftp aws-http prefetch" >> download_cmds.txt
done

parallel --jobs 10 --halt soon,fail=1 --joblog parallel_joblog_king.txt < download.sh 2> parallel_errors_king.log

```
parallel --jobs 10 --halt soon,fail=1 --joblog parallel_joblog.txt < download.sh 2> parallel_errors.log
```


## Run Sylph

```
sylph query gtdb_ordered_99.syldb -u --read-seq-id 99.9 -t 100  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./bjork_2024_query_gtdb_220_id99.tsv ./ERR*[0-9][0-9].fastq.gz
sylph profile /data1/gerryt/sylph_analyses/gtdb_98_ordered_DB/gtdb_ordered_98.syldb -u --read-seq-id 99.9 -t 100  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./bjork_2024_profile_gtdb_220_id98.tsv ./ERR*[0-9][0-9].fastq.gz
sylph profile /data1/gerryt/sylph_analyses/gtdb-r220-c200-dbv1.syldb -u --read-seq-id 99.9 -t 100  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./bjork_2024_profile_gtdb_220_id95.tsv ./ERR*[0-9][0-9].fastq.gz
```



```{r}
library(tidyverse)
library(data.table)
library(ggbeeswarm)
library(ggthemes)
```

## Load data

```{r}
long_meta <- read_csv("./data/bjork/mdat_longi_pibble.csv")
long_sprops <- read_csv("./data/bjork/sgbs_longi_pibble.csv", col_names = TRUE)
colnames(long_sprops)[[1]] <- 'sampleid'

#Check rows sum to 1, allowing for a rounding error of 0.001
stopifnot(all(abs(rowSums(long_sprops[,2:ncol(long_sprops)])-1)<0.001))

base_meta <- read_csv("./data/bjork/mdat_baseline_survival.csv")
base_sprops <- read_csv("./data/bjork/sgbs_baseline_survival.csv")
colnames(base_sprops)[[1]] <- 'sampleid'

#Check rows sum to 1, allowing for a rounding error of 0.001
stopifnot(all(abs(rowSums(base_sprops[,2:ncol(base_sprops)])-1)<0.001))
```

Melt things for plotting

```{r}
long_data <- merge(long_meta, long_sprops, by = 'sampleid', all.x = TRUE) %>% 
  as_tibble() %>%
  pivot_longer(cols = -colnames(long_meta), names_to = "species", values_to = "proportion")

long_data$SGB <- gsub("_group", "", gsub(".*t__", "", long_data$species))
```

Let's look at some of the species from the main figures in the paper.

### Figure 2

```{r}
ggplot(long_data %>% filter(SGB %in% c("SGB14993", "SGB4617")), 
       aes(x=factor(visit), y=proportion*100, col=PFS12, group=paste(visit, PFS12))) +
  geom_boxplot(outlier.color = NA) +
  geom_jitter(position = position_jitterdodge(jitter.height = 0, jitter.width = 0.2)) +
  facet_wrap(~ SGB) +
  scale_y_sqrt() +
  # scale_y_continuous(limits = c(0,5)) +
  scale_colour_manual(values = c('#8a0b80', '#d2a118')) +
  theme_bw() +
  theme(legend.position = 'none') +
  ylab("Percentage of sampled microbial diversity")
```

### Figure 3

```{r}
pdf <- long_data %>% filter(SGB %in% c("SGB4988", "SGB15124", "SGB4198", "SGB15452"))
pdf$SGB <- factor(pdf$SGB, levels = c("SGB4988", "SGB15124", "SGB4198", "SGB15452"))

ggplot(pdf, 
       aes(x=factor(visit), y=proportion*100, col=PFS12, group=paste(visit, PFS12))) +
  geom_boxplot(outlier.color = NA) +
  geom_jitter(position = position_jitterdodge(jitter.height = 0, jitter.width = 0.2)) +
  facet_wrap(~ SGB) +
  scale_y_sqrt() +
  # scale_y_continuous(limits = c(0,5)) +
  scale_colour_manual(values = c('#8a0b80', '#d2a118')) +
  theme_bw(base_size = 16) +
  theme(legend.position = 'none') +
  ylab("Percentage of sampled microbial diversity") +
  xlab("Visit")
```

## Try some classic regression analyses

```{r}
library(glmmTMB)
library(broom.mixed)

SGBs_fig13 <- c("SGB14993", "SGB4617", "SGB4988", "SGB15124", "SGB4198", "SGB15452")

beta_models <- map_dfr(SGBs_fig13, ~{
  print(.x)
  temp_data <- long_data %>% filter(SGB==.x)
  temp_data$proportion <- temp_data$proportion + 1e-8
  m1 <- glmmTMB(proportion ~  PFS12 + visit + PFS12:visit + colitis +  combi_therapy + centre + (1|patientid_anon), 
                family = beta_family(), data = temp_data)
  
  return(broom.mixed::tidy(m1) %>%
           add_column(.before=1, SGB=.x))
})

beta_models %>%
  filter(term %in% c('PFS12yes:visit')) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))
```

### Figure 4

```{r}
long_data$combination_therapy <- ifelse(long_data$combi_therapy=='no', 'Monotherapy', 'Combination')
ggplot(long_data %>% filter(SGB %in% c("SGB5121", "SGB14991", "SGB1949")), 
       aes(x=factor(visit), y=proportion*100, col=PFS12, group=paste(visit, PFS12))) +
  geom_boxplot(outlier.color = NA, position = 'dodge') +
  geom_jitter(position = position_jitterdodge(jitter.height = 0, jitter.width = 0.2)) +
  facet_grid(SGB ~ combination_therapy) +
  scale_y_sqrt() +
  # scale_y_continuous(limits = c(0,5)) +
  scale_colour_manual(values = c('#8a0b80', '#d2a118')) +
  theme_bw(base_size = 16) +
  ylab("Percentage of sampled microbial diversity") +
  xlab('visit')
```

### Let's look at total number of species as a poor proxy for sequencing depth

```{r}
long_spec_total <- long_meta
long_spec_total$nspecies <-   rowSums(long_sprops[,2:ncol(long_sprops)]>0)[match(long_spec_total$sampleid, long_sprops$sampleid)]
long_spec_total$combi_therapy <- ifelse(long_spec_total$combi_therapy=='no', 'mono', 'combination')
  
ggplot(long_spec_total, 
       aes(x=factor(visit), y=nspecies, col=PFS12, group=paste(visit, PFS12))) +
  geom_boxplot(outlier.color = NA, position = 'dodge') +
  geom_jitter(position = position_jitterdodge(jitter.height = 0, jitter.width = 0.2)) +
  facet_grid( ~ combi_therapy) +
  # scale_y_sqrt() +
  # scale_y_continuous(limits = c(0,5)) +
  scale_colour_manual(values = c('#8a0b80', '#d2a118')) +
  theme_bw() +
  ylab("Number of species") +
  xlab('visit')
```

## Try some classic regression analyses

```{r}
library(glmmTMB)
library(broom.mixed)

SGBs_fig13 <- c("SGB14993", "SGB4617", "SGB4988", "SGB15124", "SGB4198", "SGB15452")

beta_models <- map_dfr(SGBs_fig13, ~{
  print(.x)
  temp_data <- long_data %>% filter(SGB==.x)
  temp_data$proportion <- temp_data$proportion + 1e-8
  m1 <- glmmTMB(proportion ~  PFS12 + visit + PFS12:visit + colitis +  combi_therapy + centre + (1|patientid_anon), 
                family = beta_family(), data = temp_data)
  
  return(broom.mixed::tidy(m1) %>%
           add_column(.before=1, SGB=.x))
})

beta_models %>%
  filter(term %in% c('PFS12yes:visit')) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))

```


```{r}
SGBs_fig4 <- c("SGB5121", "SGB14991", "SGB1949")

beta_models <- map_dfr(SGBs_fig4, ~{
  print(.x)
  temp_data <- long_data %>% filter(SGB==.x)
  temp_data$proportion <- temp_data$proportion + 1e-8
  m1 <- glmmTMB(proportion ~  PFS12 + visit + PFS12:visit + colitis +  combi_therapy + combi_therapy:visit + 
                  centre + (1|patientid_anon), 
                family = beta_family(), data = temp_data)
  
  return(broom.mixed::tidy(m1) %>%
           add_column(.before=1, SGB=.x))
})

beta_models %>%
  filter(term %in% c('PFS12yes:visit')) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))


beta_models %>%
  filter(term %in% c('visit:combi_therapyyes')) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))


# beta_fixed_models <- map_dfr(SGBs_fig13, ~{
#   print(.x)
#   temp_data <- long_data %>% filter(SGB==.x)
#   temp_data$proportion <- temp_data$proportion + 1e-8
#   m1 <- glmmTMB(proportion ~  PFS12 + visit + PFS12:visit + colitis +  combi_therapy + centre, 
#                 family = beta_family(), data = temp_data)
#   
#   return(broom.mixed::tidy(m1) %>%
#            add_column(.before=1, SGB=.x))
# })
# 
# beta_fixed_models %>%
#   filter(term %in% c('PFS12yes:visit', 'combi_therapyyes'))
# 
# 
# logistic_models <- map_dfr(SGBs_fig13, ~{
#   print(.x)
#   temp_data <- long_data %>% filter(SGB==.x)
#   temp_data$PFS12 <- (temp_data$PFS12=="yes") * 1
#   m2 <- glmmTMB(PFS12 ~ proportion + visit + proportion:visit + colitis +  combi_therapy + centre + (1|patientid_anon), 
#                 family = binomial, data = temp_data)
#   
#   return(broom.mixed::tidy(m2) %>%
#            add_column(.before=1, SGB=.x))
# })
# 
# 
# logistic_models %>%
#   filter(term %in% c('proportion:visit', 'combi_therapyyes'))


```

```{r}

SGBs_fig13 <- c("SGB14993", "SGB4617", "SGB4988", "SGB15124", "SGB4198", "SGB15452")

beta_models <- map_dfr(SGBs_fig13, ~{
  print(.x)
  temp_data <- long_data %>% filter(SGB==.x)
  temp_data$proportion <- temp_data$proportion + 1e-8
  m1 <- glmmTMB(proportion ~  
                  visit*PFS12*combi_therapy + 
                  visit*PFS12*colitis +
                  visit*PFS12*ppi +
                  centre +
                  days_since_start_thetapy +
                  antibiotics +
                  toxicity +
                  previous_therapy +
                  age +
                  sex +
                  bmi +
                  (1|patientid_anon), 
                family = beta_family(), data = temp_data)
  
  return(broom.mixed::tidy(m1) %>%
           add_column(.before=1, SGB=.x))
})

beta_models %>%
  filter(term %in% c('PFS12yes:visit', 'combi_therapyyes'))

```


```{r}
beta_combi_models <- map_dfr(c("SGB5121", "SGB14991", "SGB1949"), ~{
  print(.x)
  temp_data <- long_data %>% filter(SGB==.x)
  temp_data$proportion <- temp_data$proportion + 1e-8
  m1 <- glmmTMB(proportion ~  PFS12 + visit + PFS12:visit + colitis +  combi_therapy:visit + centre + (1|patientid_anon), 
                family = beta_family(), data = temp_data)
  
  return(broom.mixed::tidy(m1) %>%
           add_column(.before=1, SGB=.x))
})

beta_combi_models %>%
  filter(term %in% c('PFS12yes:visit', 'visit:combi_therapyyes', 'combi_therapyyes'))

```

## A closer look at time points

```{r}
nvisits <- long_meta %>% 
  group_by(patientid_anon) %>%
  summarise(n_visits = n())

ggplot(nvisits, aes(x=n_visits)) +
  geom_histogram(binwidth = 1) +
  theme_bw(base_size = 16) +
  xlab("Number of visits") +
  ylab("Number of patients") +
  scale_x_continuous(breaks = seq(0, max(nvisits$n_visits), by=1))

long_meta %>% 
  group_by(patientid_anon, visit) %>%
  summarise(n_visits = n(),
            ndates = length(unique(date_fecal_sample))) %>%
  filter(n_visits>1) 

temp <- long_meta[long_meta$patientid_anon=="p010",]

long_meta_fixed <- long_meta %>%
  filter(patientid_anon %in% nvisits$patientid_anon[nvisits$n_visits>1]) %>%
  group_by(patientid_anon) %>%
  mutate(
    standardised_date = date_fecal_sample - min(date_fecal_sample),
    standardised_visit = order(date_fecal_sample),
    n_visits = n()
  ) %>%
  ungroup()


ggplot(long_meta_fixed, aes(x=factor(standardised_visit), y=standardised_date, col=PFS12)) +
  geom_boxplot(outlier.color = NA, position = position_dodge(width = 0.75)) +
  ggbeeswarm::geom_quasirandom(dodge.width = 0.75) +
  theme_bw(base_size = 16) +
  xlab("Standardised visit number") +
  ylab("Days from first sample") +
  scale_colour_manual(values = c('#8a0b80', '#d2a118'))

```

## StrainSpy

```{r}
library(strainspy)
library(SummarizedExperiment)
library(tidyverse)

# fread("./data/bjork/mdat_baseline_survival.csv")

sylph_98 <- read_sylph("./data/bjork/bjork_2024_profile_gtdb_220_id98.tsv.gz")

accessions <- rbind(read_tsv("./data/bjork/filereport_read_run_PRJEB70966_tsv.txt"),
      read_tsv("./data/bjork/filereport_read_run_PRJEB43119_tsv.txt"))

accessions$sample_title <- gsub("\\s.*", "", accessions$sample_title)

colData(sylph_98)$Sample_file <- gsub("_1", "", colData(sylph_98)$Sample_file)
colData(sylph_98)$sample_name <- accessions$sample_title[match(colData(sylph_98)$Sample_file, accessions$run_accession)]
colData(sylph_98)$sample_name <- gsub('-', '', colData(sylph_98)$sample_name)
colData(sylph_98)$sample_name <- gsub('BCN0', 'BCN.0', colData(sylph_98)$sample_name)
colData(sylph_98)$sample_name <- gsub('BCN1', 'BCN.1', colData(sylph_98)$sample_name)

# Still missing some
sum(!colData(sylph_98)$sample_name %in% long_meta$sampleid)
colData(sylph_98)$sample_name[!colData(sylph_98)$sample_name %in% long_meta$sampleid]
sum(!long_meta$sampleid %in% colData(sylph_98)$sample_name)
long_meta$sampleid[!long_meta$sampleid %in% colData(sylph_98)$sample_name]
long_meta[!long_meta$sampleid %in% colData(sylph_98)$sample_name,]



sylph_98 <- sylph_98[colData(sylph_98)$sample_name %in% long_meta$sampleid,]

colData(sylph_98) <- merge(colData(sylph_98), long_meta, by.x='sample_name', by.y='sampleid', all.x=TRUE)
```


```{r}
plot_pca_biplot(sylph_98, "PFS12")

plot_mds(sylph_98, "PFS12")
```


```{r}
sylph_98 <- sylph_98[,!is.na(colData(sylph_98)$PFS12)]
sylph_98 <- sylph_98[,!is.na(colData(sylph_98)$visit)]
sylph_98 <- sylph_98[,!is.na(colData(sylph_98)$colitis)]
sylph_98 <- sylph_98[,!is.na(colData(sylph_98)$combiIO)]
sylph_98 <- sylph_98[,!is.na(colData(sylph_98)$centre)]

sylph_98_filt <- strainspy::filter_by_presence(sylph_98, min_nonzero = 30)


colData(sylph_98_filt)$bmi <- scale(colData(sylph_98_filt)$bmi)
colData(sylph_98_filt)$age <- scale(colData(sylph_98_filt)$age)
colData(sylph_98_filt)$visit <- scale(colData(sylph_98_filt)$visit)

all.vars(design)[!all.vars(design) %in% colnames(colData(sylph_98_filt))]

design <- as.formula(" ~ PFS12 + visit + PFS12:visit + colitis  + antibiotics + ppi +
                  centre + (1|patientid)")
# fit <- glmZiBFit(sylph_98_filt, design, nthreads=4, method='glmmTMB')
fit <- strainspy::glmFit(sylph_98_filt, design, nthreads=1)

# design <- as.formula(" ~ PFS12 + visit + PFS12:visit + colitis +  combi_therapy +
#                 centre + gamlss::random(patientid_anon)")
# 
# fit <- glmZiBFit(sylph_98_filt, design, nthreads=4, method='gamlss')

index <- 11
colnames(fit@p_values)[index]
top_hits(fit, coef = index, alpha=0.1)


thits <- top_hits(fit, coef = "PFS12yes", alpha=0.05, method = "BH")
thits

thits_long <- top_hits(fit, coef = "PFS12yes.visit", alpha=0.05, method = "BH")
thits_long


# Issue with low sample numbers needs to be dealt with (~10 samples)
# refit <- glmZiBFit(sylph_98[rowData(sylph_98)$Genome_file %in% thits$Genome_file[1:5],], 
#           design, nthreads=1, method='glmmTMB')
# refit@convergence
# 
# refit <- glmZiBFit(sylph_98[rowData(sylph_98)$Genome_file %in% thits$Genome_file[1:5],], 
#                    design, nthreads=1, method='gamlss')
# 
# top_hits(refit, coef = 4, alpha=1)

tax <- read_taxonomy("~/Documents/strainspy_manuscript/data/bjork/sylph_DB_taxonomy_98.tsv")

thits$Species <- tax$Species[match(thits$Genome_file, tax$Genome)]

pdf <- map2_dfr(thits$Genome_file, thits$Species, ~{
  as_tibble(colData(sylph_98)) %>%
    add_column(.before=1, Genome_file = .x) %>%
    add_column(.before=1, Species = .y) %>%
    add_column(identity = as.numeric(assay(sylph_98)[rowData(sylph_98)$Genome_file==.x,]))
})


ggplot(pdf %>% filter(identity>0), aes(x=PFS12, y=identity, colour = PFS12)) +
  geom_boxplot(outlier.color = NA, position = position_dodge(width = 0.75)) +
  ggbeeswarm::geom_quasirandom(dodge.width = 0.75) +
  facet_wrap(~ Species) +
  theme_bw(base_size = 16) +
  xlab("PFS12") +
  ylab("Identity") +
  scale_colour_manual(values = c('#8a0b80', '#d2a118'))

```

```{r}
thits_long <- top_hits(fit, coef = 14, alpha=0.05)

thits_long$Species <- tax$Species[match(thits_long$Genome_file, tax$Genome)]

pdf <- map2_dfr(thits_long$Genome_file, thits_long$Species, ~{
  as_tibble(colData(sylph_98)) %>%
    add_column(.before=1, Genome_file = .x) %>%
    add_column(.before=1, Species = .y) %>%
    add_column(identity = as.numeric(assay(sylph_98)[rowData(sylph_98)$Genome_file==.x,]))
})


gg1 <- ggplot(pdf %>% filter(identity>0), aes(x=factor(visit), y=identity, colour = PFS12)) +
  geom_boxplot(outlier.color = NA, position = position_dodge(width = 1)) +
  ggbeeswarm::geom_quasirandom(dodge.width = 1) +
  facet_wrap(~ Species) +
  theme_bw(base_size = 16) +
  xlab("PFS12") +
  ylab("Identity") +
  scale_colour_manual(values = c('#8a0b80', '#d2a118'))
gg1

pdf2 <- pdf |>
  group_by(Species, PFS12, visit) |>
  summarise(presense = mean(identity>0))

gg2 <- ggplot(pdf2, aes(x=factor(visit), y=presense, fill = PFS12)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~ Species) +
  theme_bw(base_size = 16) +
  xlab("PFS12") +
  ylab("Presense") +
  scale_fill_manual(values = c('#8a0b80', '#d2a118'))
gg2

library(patchwork)
gg1 + gg2 + patchwork::plot_layout(ncol=1, guides = 'collect')


# ggplot(pdf %>% filter(Species=="COE1 sp001916965"), aes(y=identity, fill = PFS12)) +
#   geom_histogram(binwidth = 0.5) +
#   facet_grid(factor(visit) ~ PFS12) +
#   # scale_x_log10() +
#   theme_bw(base_size = 16) +
#   xlab("identity") +
#   ylab("Count") +
#   scale_fill_manual(values = c('#8a0b80', '#d2a118'))


ggplot(pdf, aes(x=factor(visit), y=identity, colour = PFS12)) +
  geom_boxplot(outlier.color = NA, position = position_dodge(width = 1)) +
  ggbeeswarm::geom_quasirandom(dodge.width = 1,  alpha=0.1) +
  facet_wrap(~ Species) +
  theme_bw(base_size = 16) +
  xlab("PFS12") +
  ylab("Identity") +
  scale_colour_manual(values = c('#8a0b80', '#d2a118'))
```

## Download data (Not available yet)

```{r}
# library(curatedMetagenomicData)
# sampleMetadata
# curatedMetagenomicData::curatedMetagenomicData()
# 
# sampleMetadata[grepl("bjork", sampleMetadata$study_name),]
```
