---
title: "Analysis of 3414 pooled colorectal cancer metagenomes"
output:
  github_document
date: "2026-04-23"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Run ZiB analysis in q99 mode

## Load dependencies

```{r, warning=FALSE, message=FALSE}
library(SummarizedExperiment)
library(tidyverse)
library(strainspy)
library(dplyr)
library(ggplot2)
library(ggvenn)
library(stringr)
library(tidyverse)
library(tidyr)
```

## Load metadata 

```{r}
meta_path <- "./data/segata_pooled_3741/combined_metadata.tsv"
meta <- read.csv(meta_path, sep = '\t') 

# set up contrasts/reference levels
meta$disease = factor(meta$disease, levels = c("Control", "CRC", "Adenoma"))
meta$study = factor(meta$study)
meta$cascon = ifelse(meta$cascon == "Case", "Case", "Control")
meta$cascon = factor(meta$cascon, levels = c("Control", "Case"))
meta$country = factor(meta$country)
meta$sex = factor(meta$sex, levels = c("Female", "Male"))
# let's reset tumour stages
meta$tumour_stage_AJCC[meta$tumour_stage_AJCC == ""] = "NoTumour"
meta$tumour_stage_AJCC[meta$disease == "Adenoma"] = "Adenoma"
meta$tumour_stage_AJCC[meta$tumour_stage_AJCC == "CRC_stage_unlabelled"] = "Unstaged"
meta$tumour_stage_AJCC = factor(meta$tumour_stage_AJCC, levels = c("NoTumour", "Adenoma", "0", "I", "II", "III", "IV", "Unstaged"))

meta$tumour_location[meta$disease == "Control"] = "NoTumour"
meta$tumour_location[meta$disease == "Adenoma"] = "Adenoma"
meta$tumour_location = factor(meta$tumour_location, levels = c("NoTumour", "Adenoma", "transverse", "left_sided", "right_sided", "multiple_sites", "nd"))

```

## Load sylph outputs

```{r}
sy <- read_sylph("./data/segata_pooled_3741/combined_q_99.tsv.gz") # q99)
sy <- filter_by_presence(sy, min_nonzero = 342) # filter at 10%

# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colnames(sy) <- gsub("_merged", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))
colData(sy)$Sample_file <- gsub("_merged", "", colnames(sy))

dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$run_accession)
all(meta$run_accession %in% colnames(sy))

sy = modify_metadata(sy, meta)
```

## Fit the model

```{r}
design <- as.formula("~ disease + age + sex + BMI + (1 | study)")

save_path <- "output_rds/CRC_zib_q_99_ebp.rds"
# Run with ebp - this looks a bit less noisy
if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit <- glmZiBFit(sy, design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit, save_path)
}
```

## Visualise Outputs

### Load GTDB taxonomy

```{r}
taxonomy <- read_taxonomy("data/TAXONOMY/sylph_DB_taxonomy_99.tsv")
```

## Manhattan and Volcano plots

### Control vs. Adenoma

```{r, fig.width=12}
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = T, coef = 3)
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = F, coef = 3)
plot_volcano(ZB_fit, coef = 3)
```

No real signal detected. This agrees with the original paper.

### Control vs. CRC

```{r, fig.width=12}
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = T, coef = 2) 
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = F, coef = 2) 
plot_volcano(ZB_fit, coef = 2)

```

Looks like there are many hits.

## Summarise and inspect

## Build a summary from top hits

```{r}
th = comp_ani_diff_and_posthoc_test(se = sy, fit = ZB_fit, th = top_hits(ZB_fit, coef = 2))
th = th[is.na(th$Comment), ]
th = strainspy:::add_tax2tophits(th, taxonomy)

# Remove the weakly separated beta hits (somewhat arbitrary 5.75% - somewhat arbitrary, but no way around manual checks to see separations
rmidx = which(abs(th$ANI_Difference) < 0.0575)
if(length(rmidx) > 0) th = th[-rmidx, ]

# Signals detected from 70 species
length(sort(table(th$Species), decreasing = T))

# Ask chatGPT for a nice summary table
summary_tbl <- th %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(
    N_hits = n(),
    
    Top_adj_p_beta  = min(p_adjust, na.rm = TRUE),
    Top_coef_beta   = coefficient[which.min(p_adjust)],
    Top_se_beta     = std_error[which.min(p_adjust)],
    Top_contig_beta = Contig_name[which.min(p_adjust)],
    
    Top_adj_p_ZI    = min(zi_p_adjust, na.rm = TRUE),
    Top_coef_ZI     = zi_coefficient[which.min(zi_p_adjust)],
    Top_se_ZI       = zi_std_error[which.min(zi_p_adjust)],
    Top_contig_ZI   = Contig_name[which.min(zi_p_adjust)],
    
    .groups = "drop"
  ) %>%
  mutate(
    Min_adj_p = pmin(Top_adj_p_beta, Top_adj_p_ZI, na.rm = TRUE),
    Dominant_component = ifelse(Top_adj_p_beta < Top_adj_p_ZI, "Beta", "ZI"),
    Dominant_contig    = ifelse(Dominant_component == "Beta", Top_contig_beta, Top_contig_ZI),
    Dominant_coef      = ifelse(Dominant_component == "Beta", Top_coef_beta, Top_coef_ZI),
    Dominant_se        = ifelse(Dominant_component == "Beta", Top_se_beta, Top_se_ZI)
  ) %>%
  select(Species, N_hits, Dominant_component, Min_adj_p, Dominant_coef, Dominant_se, Dominant_contig) %>%
  arrange(Min_adj_p)

# Write as tsv for manual perusal
# write.table(summary_tbl, "output_tables/CRC_Z99_ebp_summary.tsv", sep = '\t', col.names = T, row.names = F, quote = F)

```

## ZI component hits

When the `Dominant_component` is ZI, negative and positive `Dominant_coef` indicates presence and absence in CRC compared to control.

### Visualise this in a forest plot

```{r, paged.print=FALSE, fig.width=12, fig.height=20}
zi_summary <- summary_tbl %>%
  filter(Dominant_component == "ZI") %>%
  arrange(Min_adj_p) %>%
  mutate(
    # Confidence interval before flipping
    CI_low = Dominant_coef - 1.96 * Dominant_se,
    CI_high = Dominant_coef + 1.96 * Dominant_se,
    
    # Flip effect size and CI
    Dominant_coef = -Dominant_coef,
    CI_low = -CI_low,
    CI_high = -CI_high,
    
    # Trend annotation
    Trend = ifelse(Dominant_coef > 0, "\u2191 CRC", "\u2193 CRC")
  ) %>%
  arrange(Trend, desc(Dominant_coef)) %>%
  mutate(Species = factor(Species, levels = rev(unique(Species))))

ggplot(zi_summary, aes(x = Dominant_coef, y = Species, color = Trend)) +
  geom_point(aes(size = N_hits)) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("\u2191 CRC" = "red", "\u2193 CRC" = "blue")) +
  scale_size_continuous(
    name = "Strains",
    range = c(2, 8),  # tweak for your figure aesthetics
    breaks = c(1, 45, 90),
    labels = c("1", "45", "90"),
    guide = guide_legend(override.aes = list(linetype = 0))  # remove lines from legend
  ) +
  labs(
    x = "Effect size",
    y = "Species",
    color = "Trend"
  ) +
  theme_minimal(base_size = 16)



```

## Beta component hits

When the `Dominant_component` is Beta, negative and positive `Dominant_coef` indicates lower and higher cANI in CRC compared to control, reflecting similarity of the strain to the reference.
Higher estimates imply possibly larger strain differences in CRC compared to controls. 

```{r, paged.print=FALSE, fig.width=12, fig.height=6}

beta_summary <- summary_tbl %>%
  filter(Dominant_component == "Beta") %>%
  arrange(Min_adj_p) %>%
  mutate(
    CI_low = Dominant_coef - 1.96 * Dominant_se,
    CI_high = Dominant_coef + 1.96 * Dominant_se,
    Trend = ifelse(Dominant_coef > 0, "\u2191 CRC (similar)", "\u2193 CRC (divergent)"),
    Species = factor(Species, levels = rev(unique(Species)))
  ) %>%
  arrange(desc(Dominant_coef)) %>%
  mutate(Species = factor(Species, levels = rev(unique(Species))))

ggplot(beta_summary, aes(x = Dominant_coef, y = Species, color = Trend)) +
  geom_point(aes(size = N_hits)) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("\u2191 CRC (similar)" = "red",
                                "\u2193 CRC (divergent)" = "blue")) +
  scale_size_continuous(
    name = "Strains",
    range = c(2, 8),
    breaks = c(1, 60, 120),
    labels = c("1", "60", "120")
  ) +
  labs(
    x = "Effect size (difference in ANI, CRC vs HC)",
    y = "Species",
    color = "Trend"
  ) +
  theme_minimal(base_size = 16)
```

## ANI distribution of the top strain of each beta detected species 

```{r, fig.width=12}
plot_ani_dist(sy, phenotype = 'disease', contigs = beta_summary$Dominant_contig, 
              show_points = T, plot_type = 'box', 
              contig_names = as.character(beta_summary$Species))

th_ec = th[th$Species == "Escherichia coli",]
plot_histogram(sy, phenotype = 'disease', contig = th_ec$Contig_name[3],
              drop_zeros = T, show_quantiles = F, fit_spline = T, bins = 250)

```

# Associations with other metadata

Since `tumour_stage_AJCC` and `tumour_location` are highly correlated with `disease`, we'll first fit separate models on subsets for direct comparison.

## Tumour location 

```{r}
# Directly compare left vs. right
meta_lr_acc = which(meta$tumour_location == 'left_sided' | meta$tumour_location == 'right_sided')
lr_acc = meta$run_accession[meta_lr_acc]
meta_lr = meta[meta_lr_acc, c('run_accession', 'study', 'age', 'BMI', 'sex', 'tumour_location')]
meta_lr$tumour_location = factor(as.character(meta_lr$tumour_location), levels = c('left_sided', 'right_sided'))
# Reload sylph q99
sy <- read_sylph("./data/segata_pooled_3741/combined_q_99.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colnames(sy) <- gsub("_merged", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))
colData(sy)$Sample_file <- gsub("_merged", "", colnames(sy))

# only lVr samples
sy = sy[, sapply(lr_acc, function(x) which(x == colnames(sy)))]
sy <- filter_by_presence(sy, min_nonzero = round(dim(sy)[2]/10)) # filter at 10%

dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta_lr$run_accession)
all(meta_lr$run_accession %in% colnames(sy))

sy = modify_metadata(sy, meta_lr)

design <- as.formula("~ tumour_location + age + sex + BMI + (1 | study)")

save_path <- "output_rds/CRC_zib_q_99_ebp_tumour_location_lVr.rds"
# Run with ebp - this looks a bit less noisy
if(file.exists(save_path)){
  ZB_fit_tl <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_tl <- glmZiBFit(sy,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_tl, save_path)
}

```

### Build a summary from top hits

```{r}
th_tl = top_hits(ZB_fit_tl, coef = 2)
# There are no strain identity differences, post hoc testing ignored

th_tl$Species = taxonomy$Species[match(th_tl$Genome_file, taxonomy$Genome)]

# Signals detected from 11 species
length(sort(table(th_tl$Species), decreasing = T))

# Ask chatGPT for a nice summary table
summary_tbl <- th_tl %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(
    N_hits = n(),
    
    Top_adj_p_beta  = min(p_adjust, na.rm = TRUE),
    Top_coef_beta   = coefficient[which.min(p_adjust)],
    Top_se_beta     = std_error[which.min(p_adjust)],
    Top_contig_beta = Contig_name[which.min(p_adjust)],
    
    Top_adj_p_ZI    = min(zi_p_adjust, na.rm = TRUE),
    Top_coef_ZI     = zi_coefficient[which.min(zi_p_adjust)],
    Top_se_ZI       = zi_std_error[which.min(zi_p_adjust)],
    Top_contig_ZI   = Contig_name[which.min(zi_p_adjust)],
    
    .groups = "drop"
  ) %>%
  mutate(
    Min_adj_p = pmin(Top_adj_p_beta, Top_adj_p_ZI, na.rm = TRUE),
    Dominant_component = ifelse(Top_adj_p_beta < Top_adj_p_ZI, "Beta", "ZI"),
    Dominant_contig    = ifelse(Dominant_component == "Beta", Top_contig_beta, Top_contig_ZI),
    Dominant_coef      = ifelse(Dominant_component == "Beta", Top_coef_beta, Top_coef_ZI),
    Dominant_se        = ifelse(Dominant_component == "Beta", Top_se_beta, Top_se_ZI)
  ) %>%
  select(Species, N_hits, Dominant_component, Min_adj_p, Dominant_coef, Dominant_se, Dominant_contig) %>%
  arrange(Min_adj_p)

# Write as tsv for manual perusal
# write.table(summary_tbl, "output_tables/CRC_Z99_tVl_direct_summary.tsv", sep = '\t', col.names = T, row.names = F, quote = F)

```

There are only presence/absence hits.

### ZI component hits

Negative and positive `Dominant_coef` indicates strain presence in the guts of patients with Left and Right CRC tumors, respectively.

### Visualise this in a forest plot

```{r, paged.print=FALSE, fig.width=12, fig.height=8}
zi_summary <- summary_tbl %>%
  filter(Dominant_component == "ZI") %>%
  arrange(Min_adj_p) %>%
  mutate(
    # Confidence interval before flipping
    CI_low = Dominant_coef - 1.96 * Dominant_se,
    CI_high = Dominant_coef + 1.96 * Dominant_se,
    
    # Flip effect size and CI
    Dominant_coef = -Dominant_coef,
    CI_low = -CI_low,
    CI_high = -CI_high,
    
    # Trend annotation
    Trend = ifelse(Dominant_coef > 0, "\u2191 Right", "\u2193 Left")
  ) %>%
  arrange(Trend, desc(Dominant_coef)) %>%
  mutate(Species = factor(Species, levels = rev(unique(Species))))

ggplot(zi_summary, aes(x = Dominant_coef, y = Species, color = Trend)) +
  geom_point(aes(size = N_hits)) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("\u2191 Right" = "orange", "\u2193 Left" = "violet")) +
  scale_size_continuous(
    name = "Strains",
    range = c(2, 8),  # tweak for your figure aesthetics
    breaks = c(1, 45, 90),
    labels = c("1", "45", "90"),
    guide = guide_legend(override.aes = list(linetype = 0))  # remove lines from legend
  ) +
  labs(
    x = "Effect size",
    y = "Species",
    color = "Trend"
  ) +
  theme_minimal(base_size = 16)



```

Looks like it is mostly dominated by *Veillonella sp.* and *Pullichristensenella sp.*. 
*Veillonella parvula* is one of the top hits in 10.1016/j.chom.2025.03.012, but we are not detecting any of the other hits.

## Cancer stage

```{r, paged.print=FALSE, fig.width=12, fig.height=8}
table(meta$tumour_stage_AJCC)

# Directly compare left vs. right
meta_stage_acc = which(meta$tumour_stage_AJCC == '0' | 
                         meta$tumour_stage_AJCC == 'I' | 
                         meta$tumour_stage_AJCC == 'II' | 
                         meta$tumour_stage_AJCC == 'III' | 
                         meta$tumour_stage_AJCC == 'IV')

stage_acc = meta$run_accession[meta_stage_acc]
meta_stage = meta[meta_stage_acc, c('run_accession', 'study', 'age', 'BMI', 'sex', 'tumour_stage_AJCC')]
meta_stage$tumour_stage_AJCC = factor(as.character(meta_stage$tumour_stage_AJCC), levels = c('0', 'I', 'II', 'III', 'IV'))

###################
##### No HITS #####
###################
#### Model 1 - compare stages 1,2,3 and 4 vs. stage 0
# Reload sylph q99
sy <- read_sylph("./data/segata_pooled_3741/combined_q_99.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colnames(sy) <- gsub("_merged", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))
colData(sy)$Sample_file <- gsub("_merged", "", colnames(sy))

# only lVr samples
sy = sy[, sapply(meta_stage$run_accession, function(x) which(x == colnames(sy)))]
sy <- filter_by_presence(sy, min_nonzero = round(dim(sy)[2]/10)) # filter at 10%

dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta_stage$run_accession)
all(meta_stage$run_accession %in% colnames(sy))

sy = modify_metadata(sy, meta_stage)


save_path <- "output_rds/CRC_zib_q_99_ebp_tumour_stage_0v.rds"
design <- as.formula('~tumour_stage_AJCC + age + sex + BMI + (1 | study)')
# Run with ebp
if(file.exists(save_path)){
  ZB_fit_stage_full <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_stage_full <- glmZiBFit(sy,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_stage_full, save_path)
}

# Variables
top_hits(ZB_fit_stage_full, coef = 2, method = "BH")
top_hits(ZB_fit_stage_full, coef = 3, method = "BH")
top_hits(ZB_fit_stage_full, coef = 4, method = "BH")
top_hits(ZB_fit_stage_full, coef = 5, method = "BH")

#### Model 2 - merge stages 0 and 1 as early and compare with 2, 3 and 4.
###################
##### HAS HITS ####
###################
meta_stage$tumour_stage_merged = sapply(as.character(meta_stage$tumour_stage_AJCC), function(x) ifelse( (x=='0'|x=='I'), 'Early', x)) 
meta_stage$tumour_stage_merged = factor(as.character(meta_stage$tumour_stage_merged), levels = c('Early', 'II', 'III', 'IV'))

save_path <- "output_rds/CRC_zib_q_99_ebp_tumour_stage_early0Iv.rds"
design <- as.formula('~tumour_stage_merged + age + sex + BMI + (1 | study)')
sy = modify_metadata(sy, meta_stage)
# Run with ebp
if(file.exists(save_path)){
  ZB_fit_stage_early012V <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_stage_early012V <- glmZiBFit(sy,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_stage_early012V, save_path)
}

# Variables
th_stage_1 = cbind(top_hits(ZB_fit_stage_early012V, coef = 2, method = "BH"), model = 'early01V', stage = 2)
top_hits(ZB_fit_stage_early012V, coef = 3, method = "BH")
th_stage_2 = cbind(top_hits(ZB_fit_stage_early012V, coef = 4, method = "BH"), model = 'early01V', stage = 4)

#### No results - Fig 3B from Segata paper - merge 0, 1 and 2 as early and compare with merged 3 and 4 (late)
meta_stage$tumour_stage_merged_earlyVlate = sapply(as.character(meta_stage$tumour_stage_AJCC), function(x) ifelse( (x=='0'|x=='I'|x=='II'), 'Early', 'Late')) 
meta_stage$tumour_stage_merged_earlyVlate = factor(as.character(meta_stage$tumour_stage_merged_earlyVlate), levels = c('Early', 'Late'))

save_path <- "output_rds/CRC_zib_q_99_ebp_tumour_stage_early012Vlate34.rds"
design <- as.formula('~tumour_stage_merged_earlyVlate + age + sex + BMI + (1 | study)')
sy = modify_metadata(sy, meta_stage)
# Run with ebp
if(file.exists(save_path)){
  ZB_fit_stage_early012Vlate34 <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_stage_early012Vlate34 <- glmZiBFit(sy,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_stage_early012Vlate34, save_path)
}

# Variables
top_hits(ZB_fit_stage_early012Vlate34, coef = 2, method = "BH")

#### Has results - Fig 3C from Segata paper - merge 0, 1, 2, 3 as early and compare with 4
meta_stage$tumour_stage_merged_early3Vlate = sapply(as.character(meta_stage$tumour_stage_AJCC), function(x) ifelse( (x=='0'|x=='I'|x=='II'|x=="III"), 'Early', x)) 
meta_stage$tumour_stage_merged_early3Vlate = factor(as.character(meta_stage$tumour_stage_merged_early3Vlate), levels = c('Early', 'IV'))
sy = modify_metadata(sy, meta_stage)

save_path <- "output_rds/CRC_zib_q_99_ebp_tumour_stage_early0123V4.rds"
design <- as.formula('~tumour_stage_merged_early3Vlate + age + sex + BMI + (1 | study)')

# Run with ebp
if(file.exists(save_path)){
  ZB_fit_stage_early0123V4 <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_stage_early0123V4 <- glmZiBFit(sy,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_stage_early0123V4, save_path)
}


# Variables
th_stage_3 = cbind(top_hits(ZB_fit_stage_early0123V4, coef = 2, method = "BH"), model = 'early0123V', stage = 4)

th_stage = rbind(th_stage_1, th_stage_2, th_stage_3)
th_stage$Species = taxonomy$Species[match(th_stage$Genome_file, taxonomy$Genome)]


summary_tbl <- th_stage %>%
  filter(!is.na(Species)) %>%
  group_by(Species, stage) %>%
  summarise(
    N_hits = n(), 
    
    Top_adj_p_beta  = min(p_adjust, na.rm = TRUE),
    Top_coef_beta   = coefficient[which.min(p_adjust)],
    Top_se_beta     = std_error[which.min(p_adjust)],
    Top_contig_beta = Contig_name[which.min(p_adjust)],
    
    Top_adj_p_ZI    = min(zi_p_adjust, na.rm = TRUE),
    Top_coef_ZI     = zi_coefficient[which.min(zi_p_adjust)],
    Top_se_ZI       = zi_std_error[which.min(zi_p_adjust)],
    Top_contig_ZI   = Contig_name[which.min(zi_p_adjust)],
    
    .groups = "drop"
  ) %>%
  mutate(
    stage = stage,
    Min_adj_p = pmin(Top_adj_p_beta, Top_adj_p_ZI, na.rm = TRUE),
    Dominant_component = ifelse(Top_adj_p_beta < Top_adj_p_ZI, "Beta", "ZI"),
    Dominant_contig    = ifelse(Dominant_component == "Beta", Top_contig_beta, Top_contig_ZI),
    Dominant_coef      = ifelse(Dominant_component == "Beta", Top_coef_beta, Top_coef_ZI),
    Dominant_se        = ifelse(Dominant_component == "Beta", Top_se_beta, Top_se_ZI)
  ) %>%
  select(Species, N_hits, stage, Dominant_component, Min_adj_p, Dominant_coef, Dominant_se, Dominant_contig) %>%
  arrange(Min_adj_p)

# Write as tsv for manual perusal
# write.table(summary_tbl, "output_tables/CRC_Z99_ebp_stage_hits_summary.tsv", sep = '\t', col.names = T, row.names = F, quote = F)

plot_ani_dist(sy, phenotype = 'tumour_stage_merged_early3Vlate', contigs = summary_tbl$Dominant_contig, plot_type = 'box', show_points = F, contig_names = strainspy:::clean_contig_names(summary_tbl$Species))


```

In short, it looks like species such as *Fusobacterium animalis*, *Peptostreptococcus stomatis* and *Allisonella pneumosintes* grow in prevalence as CRC progresses from very early stages and remains stable. Beneficial species such as *Agathobacter faecis* seems to get replaced by other different strains at later stages - possibly consistent with treatment intensity.


## Paper figures
```{r}

# For CRC ZI signals, let's show the presence absence (<95% ANI) as a tile plot
zi_summary$Dominant_contig[order(abs(zi_summary$Dominant_coef), decreasing = T)[1:5]]

# Which ones to use:
# Dialister pneumosintes - NZ_CP017037.1
# Peptostreptococcus stomatis - NZ_ADGQ01000074.1
# Gemella morbillorum - NZ_LS483440.1
# Parvimonas micra - NZ_LR134472.1
# Fusobacterium animalis - NZ_CP071098.1

# Roseburia rectibacter - NZ_CP092473.1

zib_contigs = c("NZ_CP017037.1", "NZ_ADGQ01000074.1", "NZ_LS483440.1", "NZ_LR134472.1", "NZ_CP071098.1", 
                "CAKRCB010000001.1", "CABUSM010000001.1", "CAKRGY010000001.1", "CAUDRE010000001.1", "NZ_QUJS01000001.1")
sy_ss = sy[sapply(zib_contigs, function(x) grep(x, rownames(sy))),]

prev_ = list()
for(i in c("CRC", "Control")){
  sy_ss_t = as.matrix(assay(subset(sy_ss, select = colData(sy_ss)$disease  %in% i)))
  prev_[[i]] = apply(sy_ss_t, 1, function(x) sum(x >= 95)/ncol(sy_ss_t))
}
prev_df = data.frame(do.call(cbind, prev_))
prev_df$Species = c("Dialister pneumosintes", 
                    "Peptostreptococcus stomatis",
                    "Gemella morbillorum",
                    "Parvimonas micra",
                    "Fusobacterium animalis",
                    "Lachnospira sp.",
                    "Faecalibacterium sp.",
                    "Butyribacter hominis",
                    "Faecalibacterium prausnitzii",
                    "Butyribacter intestini")


prev_df <- prev_df %>%
  mutate(
    log2FC = log2(CRC / Control),
    direction = ifelse(log2FC > 0, "Enriched in CRC", "Depleted in CRC")
  )

# explicit ordering: CRC group first (high to low), then Control group
crc_df <- prev_df %>%
  filter(direction == "Enriched in CRC") %>%
  arrange(desc(CRC))

ctrl_df <- prev_df %>%
  filter(direction == "Depleted in CRC") %>%
  arrange(CRC)

prev_df$Species <- factor(
  prev_df$Species,
  levels = c(ctrl_df$Species, crc_df$Species)
)

ggplot(prev_df, aes(y = Species)) +
  
  geom_segment(
    aes(x = Control, xend = CRC, yend = Species, color = direction),
    linewidth = 1.2
  ) +
  
  geom_point(aes(x = Control, color = direction),
             shape = 21, fill = "white", size = 3) +
  
  geom_point(aes(x = CRC, color = direction),
             shape = 21, fill = "black", size = 3) +
  
  scale_color_manual(values = c(
    "Enriched in CRC" = "#d73027",
    "Depleted in CRC" = "#4575b4"
  )) +
  
  labs(x = "Prevalence", y = NULL, color = NULL) +
  
  theme_classic(base_size = 16) +   # 👈 main change
  
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 13),
    axis.title.x = element_text(size = 15, face = "bold"),
    
    legend.position = "top",
    legend.text = element_text(size = 13),
    
    axis.line = element_line(linewidth = 0.7)
  )

#####################################
do_lolipop = function(th_sp){
  
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  
  ## =========================================================
  ## species name
  ## =========================================================
  
  species_name = unique(th_sp$Species)
  
  ## =========================================================
  ## get submatrices
  ## =========================================================
  
  sy_ec = sy[
    sapply(
      th_sp$Contig_name,
      function(x) which(rownames(sy) %in% x)
    ),
  ]
  
  sy_ec_crc = as.matrix(
    assay(
      subset(
        sy_ec,
        select = colData(sy_ec)$disease %in% "CRC"
      )
    )
  )
  
  sy_ec_ctr = as.matrix(
    assay(
      subset(
        sy_ec,
        select = colData(sy_ec)$disease %in% "Control"
      )
    )
  )
  
  ## =========================================================
  ## mean ANI per strain
  ## =========================================================
  
  crc_mean <- apply(
    sy_ec_crc,
    1,
    function(x)
      mean(x[x > 0], na.rm = TRUE)
  )
  
  ## reconstruct HC from model-estimated contrast
  hc_mean <- crc_mean - th_sp$ANI_Difference
  
  ## =========================================================
  ## plotting dataframe
  ## =========================================================
  
  df_plot <- data.frame(
    strain = names(hc_mean),
    HC = hc_mean,
    CRC = crc_mean,
    effect = th_sp$ANI_Difference,
    padj = th_sp$p_adjust
  )
  
  ## direction
  df_plot$direction <- df_plot$CRC > df_plot$HC
  
  ## WLS weights
  df_plot$weights <- -log10(df_plot$padj + 1e-10)
  
  ## avoid zero / inf weirdness
  df_plot$weights[!is.finite(df_plot$weights)] <- max(
    df_plot$weights[is.finite(df_plot$weights)],
    na.rm = TRUE
  )
  
  ## =========================================================
  ## long format
  ## =========================================================
  
  plot_long <- df_plot %>%
    pivot_longer(
      cols = c(HC, CRC),
      names_to = "group",
      values_to = "ANI"
    ) %>%
    left_join(
      df_plot %>%
        select(
          strain,
          direction,
          weights
        ),
      by = "strain"
    )
  
  ## =========================================================
  ## weighted overall ANI shift
  ## =========================================================
  
  overall_shift <- weighted.mean(
    df_plot$effect,
    w = df_plot$weights,
    na.rm = TRUE
  )
  
  ## =========================================================
  ## plot
  ## =========================================================
  cat(min(plot_long$ANI))
  cat(max(plot_long$ANI))
  
  p = ggplot(
    plot_long,
    aes(
      x = group,
      y = ANI,
      group = strain
    )
  ) +
    
    ## strain trajectories
    geom_line(
      aes(
        colour = direction.x
      ),
      alpha = 0.55,
      linewidth = 0.7
    ) +
    
    ## endpoints
    geom_point(
      size = 1.8,
      colour = "black",
      alpha = 0.8
    )
  
  ## only add trend if >1 strain
  if(nrow(th_sp) > 1){
    
    p = p +
      
      geom_smooth(
        aes(
          group = 1,
          weight = weights.x
        ),
        method = "lm",
        se = FALSE,
        linewidth = 1.2,
        linetype = "dashed",
        colour = "grey30"
      )
  }
  
  p = p +
    
    scale_colour_manual(
      values = c(
        "TRUE"  = "#d95f5f",
        "FALSE" = "#5f8fd9"
      ),
      guide = "none"
    ) +
    
    coord_cartesian(
      ylim = c(96.5, 98.75)
    ) +
    
    labs(
      title = species_name,
      subtitle = paste0(
        "Weighted mean ΔANI (CRC − HC) = ",
        round(overall_shift, 4)
      ),
      x = NULL,
      y = "Mean ANI"
    ) +
    
    theme_classic() +
    
    theme(
      
      plot.title = element_text(
        face = "italic",
        size = 16
      ),
      
      plot.subtitle = element_text(
        size = 16,
        colour = "grey30"
      ),
      
      axis.text = element_text(
        size = 16
      ),
      
      axis.title.y = element_text(
        size = 16
      ),
      
      legend.position = "none"
    )
  
  return(p)
}

th_beta = th[th$hit_component=="beta",]
th_beta = th_beta[order(abs(th_beta$ANI_Difference), decreasing = T),]

# species_name <- "Roseburia sp019411365"
# th_sp <- th_beta %>%
#   filter(Species == species_name)
# do_lolipop(th_sp)

# These effect sizes are too small
# species_name <- "Blautia_A wexlerae"
# th_sp <- th_beta %>%
#   filter(Species == species_name)
# do_lolipop(th_sp)
# 
# 
# species_name <- "Fusicatenibacter saccharivorans"
# th_sp <- th_beta %>%
#   filter(Species == species_name)
# do_lolipop(th_sp)

# Just a single strain, probably not good coverage in GTDB
# species_name <- "Bacteroides fragilis"
# th_sp <- th_beta %>%
#   filter(Species == species_name)
# do_lolipop(th_sp)

# Two opportunists
# Pick this one, 0.07 difference!

species_name <- "Escherichia coli"
th_sp <- th_beta %>%
  filter(Species == species_name)
p1 = do_lolipop(th_sp)

species_name <- "Streptococcus vestibularis"
th_sp <- th_beta %>%
  filter(Species == species_name)
p2 = do_lolipop(th_sp)

# Two good ones
species_name <- "Faecalibacterium prausnitzii_D"
th_sp <- th_beta %>%
  filter(Species == species_name)
p3 = do_lolipop(th_sp)

species_name <- "Faecalibacterium longum"
th_sp <- th_beta %>%
  filter(Species == species_name)
p4 = do_lolipop(th_sp)

p1 + p2 + p3 + p4



```
