---
title: "Prediction of CRC using 3414 pooled colorectal cancer metagenomes"
output:
  github_document
date: "2025-11-01"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up prediction model

## Load deps
```{r, warning=FALSE, message=FALSE}
library(strainspy)
library(SummarizedExperiment)
library(caret)
library(glmnet) # elastic net
library(ranger) # RF
library(pROC)
library(doParallel)
library(foreach)
library(parallel)
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Using query99 data and fit

```{r}
meta_path <- "data/segata_pooled_3741/combined_metadata.tsv"
meta <- read.csv(meta_path, sep = '\t')

# set up contrasts/reference levels
meta$disease = factor(meta$disease, levels = c("Control", "CRC", "Adenoma"))
meta$study = factor(meta$study)
meta$cascon = ifelse(meta$cascon == "Case", "Case", "Control")
meta$cascon = factor(meta$cascon, levels = c("Control", "Case"))
meta$country = factor(meta$country)
meta$sex = factor(meta$sex, levels = c("Female", "Male"))
# let's reset tumour stages
meta$tumour_stage_AJCC[meta$tumour_stage_AJCC == ""] = "NoTumour"
meta$tumour_stage_AJCC[meta$disease == "Adenoma"] = "Adenoma"
meta$tumour_stage_AJCC[meta$tumour_stage_AJCC == "CRC_stage_unlabelled"] = "Unstaged"
meta$tumour_stage_AJCC = factor(meta$tumour_stage_AJCC, levels = c("NoTumour", "Adenoma", "0", "I", "II", "III", "IV", "Unstaged"))

meta$tumour_location[meta$disease == "Control"] = "NoTumour"
meta$tumour_location[meta$disease == "Adenoma"] = "Adenoma"
meta$tumour_location = factor(meta$tumour_location, levels = c("NoTumour", "Adenoma", "transverse", "left_sided", "right_sided", "multiple_sites", "nd"))

sy <- read_sylph("data/segata_pooled_3741/combined_q_99.tsv.gz") # q99)
sy <- filter_by_presence(sy, min_nonzero = 342) # filter at 10%

# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colnames(sy) <- gsub("_merged", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))
colData(sy)$Sample_file <- gsub("_merged", "", colnames(sy))

dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$run_accession)
all(meta$run_accession %in% colnames(sy))


# Merge adenoma and HC into None 
meta$disease = ifelse(meta$disease == "CRC", "CRC", "None")

sy = modify_metadata(sy, meta, replace = T)

# Read in the fit from save path
fit = readRDS("output_rds/CRC_zib_q_99_ebp.rds")
```


## Try a few models

### Elastic net using FDR predictors

```{r}
th = top_hits(fit, coef = 2, method = "BH")

sy_mx_BH = strainspy::prep_for_prediction(sy, 'disease', th$Contig_name)

# sy_mx = assay(sy)
# sy_mx = as.matrix(t(sy_mx[rownames(sy_mx) %in% th$Contig_name, ]))
# 
# sy_mx = sy_mx[,order(match(colnames(sy_mx), th$Contig_name))]
# 
# all(colnames(sy_mx) == th$Contig_name)
# 
# # These long column names are annoying and R will likely not accept it, I'll change to Genome names
# colnames(sy_mx) = th$Genome_file
# 
# sy_mx_BH = cbind(disease = as.factor(ifelse(meta$disease == "CRC", "CRC", "None")),
#               as.data.frame(sy_mx))

save_path = "output_rds/pred_enet_crc_q99_BH.rds"

if(file.exists(save_path)) {
  enet_fit_BH = readRDS(save_path)
} else {
  enet_fit_BH <- caret::train(disease ~ .,
                              data = sy_mx_BH,
                              method = 'glmnet',
                              preProcess = c("center", "scale"),
                              metric = "ROC",
                              trControl = trainControl(
                                method = "cv",
                                classProbs = TRUE,              # allow probabilities
                                summaryFunction = twoClassSummary,  # compute ROC, Sens, Spec
                                index = groupKFold(meta$study, k = length(table(meta$study))),
                                savePredictions = "final"
                              ),
                              weights = ifelse(sy_mx_BH$disease=="CRC", 2, 1))
  
  saveRDS(enet_fit_BH, save_path)
}
```

### Random Forest using FDR predictors

```{r}

save_path = "output_rds/pred_rf_crc_q99_BH.rds"

if(file.exists(save_path)) {
  rf_fit_BH = readRDS(save_path)
} else {
  rf_fit_BH <- caret::train(disease ~ .,
                            data = sy_mx_BH,
                            method = 'ranger',
                            preProcess = c("center", "scale"),
                            metric = "ROC",
                            trControl = trainControl(
                              method = "cv",
                              classProbs = TRUE,              # allow probabilities
                              summaryFunction = twoClassSummary,  # compute ROC, Sens, Spec
                              index = groupKFold(meta$study, k = length(table(meta$study))),
                              savePredictions = "final"
                            ),
                            weights = ifelse(sy_mx_BH$disease=="CRC", 2, 1))
  
  saveRDS(rf_fit_BH, save_path)
}


```

### Elastic net using Bonferroni predictors at stricter alpha

```{r}

# With Bonferroni - restricting to the most informative predictors
th = top_hits(fit, coef = 2, method = "bonferroni", alpha = 0.01)

sy_mx_bf = strainspy::prep_for_prediction(sy, 'disease', th$Contig_name)

# sy_mx = assay(sy)
# sy_mx = as.matrix(t(sy_mx[rownames(sy_mx) %in% th$Contig_name, ]))
# 
# sy_mx = sy_mx[,order(match(colnames(sy_mx), th$Contig_name))]
# 
# all(colnames(sy_mx) == th$Contig_name)
# 
# # These long column names are annoying and R will likely not accept it, I'll change to Genome names
# colnames(sy_mx) = th$Genome_file
# 
# sy_mx_bf = cbind(disease = as.factor(ifelse(meta$disease == "CRC", "CRC", "None")),
#               as.data.frame(sy_mx))
save_path = "output_rds/pred_enet_crc_q99_bf.rds"

if(file.exists(save_path)) {
  enet_fit_bf = readRDS(save_path)
} else {
  enet_fit_bf <- caret::train(disease ~ .,
                              data = sy_mx_bf,
                              method = 'glmnet',
                              preProcess = c("center", "scale"),
                              metric = "ROC",
                              trControl = trainControl(
                                method = "cv",
                                classProbs = TRUE,              # allow probabilities
                                summaryFunction = twoClassSummary,  # compute ROC, Sens, Spec
                                index = groupKFold(meta$study, k = length(table(meta$study))),
                                savePredictions = "final"
                              ),
                              weights = ifelse(sy_mx_bf$disease=="CRC", 2, 1))
  
  
  saveRDS(enet_fit_bf, save_path)
}

```

## Visualise all this as ROC and AUC
```{r, fig.width=12}
# Do the predictions
pred_BH <- predict(enet_fit_BH, sy_mx_BH, type = "prob")$CRC
pred_Bonf <- predict(enet_fit_bf, sy_mx_bf, type = "prob")$CRC
pred_BH_rf <- predict(rf_fit_BH, sy_mx_BH, type = "prob")$CRC

best_BH <- enet_fit_BH$bestTune
cv_BH <- subset(enet_fit_BH$pred, alpha==best_BH$alpha & lambda==best_BH$lambda)

best_Bonf <- enet_fit_bf$bestTune
cv_Bonf <- subset(enet_fit_bf$pred, alpha==best_Bonf$alpha & lambda==best_Bonf$lambda)

best_BH_RF <- rf_fit_BH$bestTune
cv_BH_RF <- subset(rf_fit_BH$pred, mtry==best_BH_RF$mtry & min.node.size==best_BH_RF$min.node.size)

roc_BH_cv <- roc(cv_BH$obs, cv_BH$CRC, levels=c("None","CRC"))
roc_Bonf_cv <- roc(cv_Bonf$obs, cv_Bonf$CRC, levels=c("None","CRC"))
roc_BH_RF_cv <- roc(cv_BH_RF$obs, cv_BH_RF$CRC, levels=c("None","CRC"))

roc_BH <- roc(sy_mx_BH$disease, pred_BH, levels = c("None","CRC"))
roc_Bonf <- roc(sy_mx_bf$disease, pred_Bonf, levels = c("None","CRC"))
roc_BH_RF <- roc(sy_mx_bf$disease, pred_BH_rf, levels = c("None","CRC"))

# roc_BH_cv <- roc(cv_BH$obs, cv_BH$CRC, levels=c("None","CRC"))
# roc_Bonf_cv <- roc(cv_Bonf$obs, cv_Bonf$CRC, levels=c("None","CRC"))


plot(roc_BH_RF, col="darkgreen", lwd=2)
plot(roc_BH, col="blue", lwd=2, main="ROC Comparison", add=TRUE)
plot(roc_Bonf, col="red", lwd=2, add=TRUE)

plot(roc_BH_RF_cv, col="green", lwd=2, lty=2, add=TRUE)
plot(roc_BH_cv, col="cyan", lwd=2, lty=2, add=TRUE)
plot(roc_Bonf_cv, col="magenta", lwd=2, lty=2, add=TRUE)


legend("bottomright",
       legend=c(
         paste0("RF FDR BH: AUC=", round(auc(roc_BH_RF), 3)),
         paste0("ENET FDR BH: AUC=", round(auc(roc_BH), 3)),
         paste0("ENET Bonferroni: AUC=", round(auc(roc_Bonf), 3)),
         
         paste0("RD FDR BH CV: AUC=", round(auc(roc_BH_RF_cv), 3)),
         paste0("ENET FDR BH CV: AUC=", round(auc(roc_BH_cv), 3)),
         paste0("ENET Bonferroni CV: AUC=", round(auc(roc_Bonf_cv), 3))
         
       ),
       col=c("blue","red","darkgreen", "cyan","magenta", "green"),
       lwd=2, lty=c(1,1,1,22,2))


```

Looks like RF full model is overfitting, but this is expected to an extent.

Bonferroni predictors lose sensitivity, we'll use FDR (BH).

# LODO 

## Using Elastic net

### Set up pipeline
```{r}

# Try to parallelise
# registerDoParallel(cores = parallel::detectCores() - 1)

# Let's do by meta$study
meta = meta[match(colnames(sy), meta$run_accession), ]
all(meta$run_accession == colnames(sy)) # order looks okay

dset = as.character(unique(meta$study))

# There is no point in fitting the whole model for each training set, we'll use the global lot of predictors
# We are biasing things a bit, but probably okay?

th = top_hits(fit, coef = 2, method = "BH")
```

### Estimate AUC
```{r}
AUCs_enet =  list()
for(d in dset){
  
  hold_out_idx = which(meta$study == d)
  hold_out_meta = meta[hold_out_idx, ]
  hold_out_sy = sy[ ,hold_out_idx]
  
  train_idx = which(meta$study != d)
  train_meta = meta[train_idx, ]
  train_sy = sy[ ,train_idx]
  
  save_path = paste("output_rds/pred_enet_crc_q99_BH_lodo_", d ,".rds", collapse = "", sep = "")
  
  if(file.exists(save_path)) {
    enet_fit_lodo = readRDS(save_path)
  } else {
    
    train_mx = strainspy::prep_for_prediction(train_sy, 'disease', th$Contig_name)
    
    enet_fit_lodo <- caret::train(
      disease ~ .,
      data = train_mx,
      method = 'glmnet',
      preProcess = c("center", "scale"),
      metric = "ROC",
      trControl = caret::trainControl(
        method = "cv",
        classProbs = TRUE,
        summaryFunction = twoClassSummary,
        index = groupKFold(train_meta$study, k = length(unique(train_meta$study))),
        savePredictions = "final",
        allowParallel = TRUE   # <- important
      ),
      weights = ifelse(train_mx$disease=="CRC", 2, 1))
    
    
    saveRDS(enet_fit_lodo, save_path)
  }
  
  hold_out_mx = strainspy::prep_for_prediction(hold_out_sy, 'disease', th$Contig_name)
  
  if(length(unique(hold_out_mx$disease))  == 2){
    
    pred_lodo <- predict(enet_fit_lodo, hold_out_mx, type = "prob")$CRC
    roc_lodo <- roc(factor(hold_out_mx$disease, levels = c("None", "CRC")), pred_lodo, levels = c("None","CRC"))
    AUCs_enet[[d]] = auc(roc_lodo)
    cat('Done with', d, 'AUC = ', AUCs_enet[[d]], '\n')
  } else {
    cat('Done with', d, 'cannot compute AUC, dataset is all:', as.character(unique(hold_out_mx$disease)), '\n')
  }
  
  
}

summary(unlist(AUCs_enet))

```

## Using RF

```{r}
AUCs_RF =  list()
for(d in dset){
  hold_out_idx = which(meta$study == d)
  hold_out_meta = meta[hold_out_idx, ]
  hold_out_sy = sy[ ,hold_out_idx]
  
  train_idx = which(meta$study != d)
  train_meta = meta[train_idx, ]
  train_sy = sy[ ,train_idx]
  
  save_path = paste("output_rds/pred_rf_crc_q99_BH_lodo_", d ,".rds", collapse = "", sep = "")
  
  if(file.exists(save_path)) {
    rf_fit_lodo = readRDS(save_path)
  } else {
    train_mx = strainspy::prep_for_prediction(train_sy, 'disease', th$Contig_name)
    
    rf_fit_lodo <- caret::train(disease ~ .,
                                data = train_mx,
                                method = 'ranger',
                                preProcess = c("center", "scale"),
                                metric = "ROC",
                                trControl = trainControl(
                                  method = "cv",
                                  classProbs = TRUE,
                                  summaryFunction = twoClassSummary,
                                  # index = groupKFold(train_meta$study, k = length(unique(train_meta$study))),
                                  savePredictions = "final",
                                  allowParallel = TRUE
                                ),
                                weights = ifelse(train_mx$disease=="CRC", 2, 1))
    
    
    saveRDS(rf_fit_lodo, save_path)
  }
  
  hold_out_mx = strainspy::prep_for_prediction(hold_out_sy, 'disease', th$Contig_name)
  
  if(length(unique(hold_out_mx$disease))  == 2){
    
    pred_lodo <- predict(rf_fit_lodo, hold_out_mx, type = "prob")$CRC
    roc_lodo <- roc(factor(hold_out_mx$disease, levels = c("None", "CRC")), pred_lodo, levels = c("None","CRC"))
    AUCs_RF[[d]] = auc(roc_lodo)
    cat('Done with', d, 'AUC = ', AUCs_RF[[d]], '\n')
  } else {
    cat('Done with', d, 'cannot compute AUC, dataset is all:', as.character(unique(hold_out_mx$disease)), '\n')
  }
  
  
}

summary(unlist(AUCs_RF))
```

## Compare with published results
```{r}
# Published results
AUCs_pub = list()
AUCs_pub[["YuJ_2015"]] = 0.93
AUCs_pub[["VogtmannE_2016"]] = 0.81
# AUCs_pub[["c2_COLOBIOME"]] = 1
AUCs_pub[["c3_IIGM_CZ"]] = 0.74
AUCs_pub[["c4_IIGM_IT"]] = 0.78
AUCs_pub[["WirbelJ_2018"]] = 0.87
AUCs_pub[["ZellerG_2014"]] = 0.83
AUCs_pub[["FengQ_2015"]] = 0.87
AUCs_pub[["YachidaS_2019"]] = 0.71
AUCs_pub[["YangJ_2020"]] = 0.93
AUCs_pub[["LiuNN_2022"]] = 0.85
AUCs_pub[["YangY_2021"]] = 0.81
AUCs_pub[["c6__IIGM_TU"]] = 0.96
# AUCs_pub[["c5_NSHII"]] = 1
AUCs_pub[["GuptaA_2019"]] = 0.97
AUCs_pub[["ThomasAM_2018b"]] = 0.89

# Take the same subset
AUCs_RF_pub = sapply(names(AUCs_pub), function(X) AUCs_RF[[X]])
AUCs_enet_pub = sapply(names(AUCs_pub), function(X) AUCs_enet[[X]])
AUCs_pub = unlist(AUCs_pub)

# Summary
summary(AUCs_pub)

# Summary
summary(AUCs_RF_pub)

# Summary
summary(AUCs_enet_pub)

plt_df = data.frame(dataset = names(AUCs_pub),
                    paper = AUCs_pub, 
                    RF = AUCs_RF_pub,
                    enet = AUCs_enet_pub)
plt_long <- plt_df %>%
  select(dataset, paper, RF, enet) %>%
  pivot_longer(cols = c(paper, RF, enet),
               names_to = "method",
               values_to = "AUC")

plt_long$method = factor(plt_long$method, levels = c("paper", "RF", "enet"))

# plot
ggplot(plt_long, aes(x = dataset, y = AUC, fill = method)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("paper" = "#4E79A7", "RF" = "#F28E2B", "enet" = "#E15759")) +
  labs(x = "Dataset", y = "AUC", fill = "Method") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    legend.position = "top"
  )

# plot
ggplot(plt_long, aes(x = dataset, y = method, fill = AUC)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = round(AUC, 2)), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#FBE9E7", high = "#D84315", name = "AUC") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 12),
    panel.grid = element_blank(),
    legend.position = "right"
  )



```