---
title: "Using gut microbial strains to predict cancer immunotherapy outcomes - a study in rare cancers and melanoma"
output:
  github_document
date: "2026-06-16"
editor_options:
  chunk_output_type: console
---
## Load dependencies

```{r, warning=FALSE, message=FALSE}
library(strainspy)
library(SummarizedExperiment)
library(caret)
library(glmnet) # elastic net
library(ggrepel)
library(ranger) # RF
library(pROC)
library(doParallel)
library(foreach)
library(parallel)
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggtree)
library(ggsci)
```

# Prediction

Instead of building a set of predictors using all genomes found through assembly, GTDB, etc. we'll use StrainSpy with GTDB reference and find a small subset of significant hits in rare cancers and use those strain ANIs to predict outcomes in Melanoma. This avoids dat lekage and attempts to recreate Gunjur et al. Fig 4d using a different approach

## Set up some functions

We will be using Elastic net for everything - less resource intensive, but probably less accurate.

```{r}
run_enet <- function(train_idx, test_idx, meta, sy, feature_names, return_fit = F) {
  
  train_mx <- strainspy::prep_for_prediction(
    sy[, train_idx],
    'RvsP',
    feature_names
  )
  
  cl <- makePSOCKcluster(detectCores() - 1)
  registerDoParallel(cl)
  
  set.seed(1988)
  fit <- caret::train(
    RvsP ~ .,
    data = train_mx,
    method = "glmnet",
    preProcess = c("center", "scale"),
    metric = "ROC",
    trControl = caret::trainControl(
      method = "cv",
      classProbs = TRUE,
      summaryFunction = twoClassSummary,
      savePredictions = "final"
    )
  )
  
  if(return_fit) {
    # If return_fit, we only use the test_idx to see if it has both classes
    return(fit)
  }
  
  
  test_mx <- strainspy::prep_for_prediction(
    sy[, test_idx],
    'RvsP',
    feature_names
  )
  
  # Skip if test labels are all one class
  if (length(unique(test_mx$RvsP)) < 2) {
    warning("Test class only has one type of label")
    return(NA)
  }
  
  
  preds <- predict(fit, test_mx, type = "prob")$R
  roc_obj <- roc(test_mx$RvsP, preds, levels = c("NR","R"))
  as.numeric(auc(roc_obj))
}

print_auc <- function(boot_auc, ret = F){
  point_est <- median(boot_auc)
  ci_lower  <- quantile(boot_auc, 0.025)
  ci_upper  <- quantile(boot_auc, 0.975)
  
  cat("AUC =", round(point_est, 3), "[95% CI:", round(ci_lower, 3), "-", round(ci_upper, 3), "]\n")
  
  if(ret){
    return(c(point_est, ci_lower, ci_upper))
  }
}

evaluate_test_set <- function(fit, test_idx, sy, th_filt_names, B = 1000) {
  if (length(test_idx) == 0) return(c(0, 0, 0))
  
  # Prepare data and predict
  test_mx <- strainspy::prep_for_prediction(sy[, test_idx, drop = FALSE], 'RvsP', th_filt_names)
  pred_probs <- predict(fit, test_mx, type = "prob")$R
  
  nr_idx <- which(test_mx$RvsP == "NR")
  r_idx  <- which(test_mx$RvsP == "R")
  
  n_nr <- length(nr_idx)
  n_r  <- length(r_idx)
  
  # Bootstrap loop
  boot_auc <- replicate(B, {
    # Sample from NR and R separately to maintain exact class proportions
    boot_nr <- sample(nr_idx, size = n_nr, replace = TRUE)
    boot_r  <- sample(r_idx,  size = n_r,  replace = TRUE)
    
    # Combine them into our bootstrap index vector
    boot_idx <- c(boot_nr, boot_r)
    
    # Compute AUC cleanly without worrying about class vanishing
    roc_obj <- roc(test_mx$RvsP[boot_idx], pred_probs[boot_idx], 
                   levels = c("NR", "R"), direction = "<", quiet = TRUE)
    as.numeric(auc(roc_obj))
  })
  
  boot_auc <- boot_auc[!is.na(boot_auc)]
  if (length(boot_auc) == 0) return(c(0, 0, 0))
  
  return(print_auc(boot_auc, ret = TRUE)) # Returns c(bmed, lci, uci)
}


make_paper_heatmap = function(Op){
  ggplot(Op, aes(Test, Train, fill = plot_value)) +
    
    geom_tile(color = "white", linewidth = 0.6) +
    
    geom_text(
      aes(label = ifelse(is.na(plot_value), "", sprintf("%.2f", plot_value))),
      size = 3.5,
      fontface = "bold"
    ) +
    
    scale_fill_gradientn(
      colours = c("grey", "grey", "#ff0000"), 
      values = scales::rescale(c(0.2, 0.55, 1)),
      limits = c(0, 1),
      name = "AUC",
      na.value = "black"
    ) +
    
    coord_fixed() +
    
    facet_grid(
      therapy_train ~ therapy_test,
      scales = "free",
      space = "free"
    ) +
    
    
    theme_bw(base_size = 14) +
    
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "grey95"),
      strip.text = element_text(face = "bold")
    )
}

```

## Classic cross prediction

```{r}
# Load the merged dataset
meta_path <- "./data/ash_pancancer/metadata_full.tsv"
meta_pan <- read.csv(meta_path, sep = '\t') 
meta_pan = cbind(run_acc = meta_pan$run_accession, meta_pan)
# From paper:
# RvsP = CR or PR VS. PD or cPD - excluded patients with a BOR of stable disease (SD) (n=29)

# Outcome
meta_pan$RvsP = "R"
meta_pan$RvsP[which(meta_pan$BOR == "PD" | meta_pan$BOR == "cPD")] = "NR" 
meta_pan$RvsP = factor(meta_pan$RvsP, levels = c("NR", "R"))

# Melanoma meta
meta_m = read.csv("data/melanoma_pooled/meta_melanoma_sm.csv")

meta = rbind(data.frame(X = meta_pan$run_acc, c_type = "RARE", type = meta_pan$histology_cohort.x, therapy_type = "combination", RvsP = meta_pan$RvsP), 
             data.frame(X = meta_m$X, c_type = "MEL", type = meta_m$Study_simplified, therapy_type = meta_m$ICB, RvsP = meta_m$ORR))
meta$therapy_type <- ifelse(meta$therapy_type %in% c("aPD1/aCTLA", "combination"), "CICB", ifelse(meta$therapy_type == "aPD1", "PD1", "other"))
meta$therapy_type[is.na(meta$therapy_type)] = "other" # Get this of the NAs 

sy <- read_sylph("./data/ash_pancancer/merged_pan_melanoma.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

# Reorder meta 
meta = meta[match(colnames(sy), meta$X), ]

# Checks before merging metadata
all(colnames(sy) %in% meta$X)
all(meta$X %in% colnames(sy))

sy = modify_metadata(sy, meta, replace = T)

# Keep only the relevant samples
keep_pan_samples = meta_pan$run_acc[meta_pan$BOR != "SD"]
keep_mela_samples = readLines("data/melanoma_pooled/gunjur_samples.txt")

sy = sy[ , colnames(sy) %in% c(keep_pan_samples, keep_mela_samples)]
sy = filter_by_presence(sy, ceiling(dim(sy)[2]/10))
dim(sy)


dset = unique(sy$type)
B = 1000
therapy_types <- c("CICB", "PD1")
design = as.formula('~ RvsP')
Op = data.frame()

for(train_dset_nm in dset){
  for (train_type in therapy_types) {
    
    train_idx <- which(sy$type == train_dset_nm & sy$therapy_type == train_type)
    
    sy_train = sy[, train_idx]
    sy_train = filter_by_presence(sy_train, ceiling(dim(sy_train)[2]/10))
    
    if(length(train_idx) > 20) { # We can train with this
      
      cat("Running strainspy for", train_type, "in", train_dset_nm, "\n")
      
      save_path = paste("output_rds/pred_enet_imm_fit_train_", train_dset_nm, "_",  train_type ,".rds", sep = "")
      if(file.exists(save_path)){
        ZB_fit = readRDS(save_path)
      } else {
        ebp = compute_eb_priors(sy_train, design = design, nthreads = parallel::detectCores(), low_cutoff = 0, high_cutoff = Inf)
        ZB_fit = glmZiBFit(sy_train, design = design, MAP_prior = ebp, nthreads = parallel::detectCores())
        
        saveRDS(ZB_fit, save_path)
      }
      
      # If the model is trained, we can test with it
      
      # Selecting features. If the fit has significant features, use those
      th = top_hits(ZB_fit)
      th = comp_ani_diff_and_posthoc_test(sy_train, ZB_fit, th)
      th = th[is.na(th$Comment),]
      
      if(nrow(th) > 2){
        feature_names = th$Contig_name[1:min(100,nrow(th))] 
      } else { # If we don't have any significant features, let's select a subset
        # cat("No singificant hits for therapy", train_type, "in", train_dset_nm, "\n")
        # next
        th = top_hits(ZB_fit, alpha = 1)[1:150, ] # This is arbitrary
        th = comp_ani_diff_and_posthoc_test(sy_train, ZB_fit, th)
        th = th[is.na(th$Comment),]
      }
      
      
      
      
      
      
      enet_mdl = run_enet(train_idx = train_idx, test_idx = "", sy = sy, meta = meta, return_fit = T, feature_names)
      
      
      for (test_dset_nm in dset) {
        for (test_type in therapy_types) {
          # Determine test indices based on type criteria (preserving your original != 'CICB' logic for 'Other')
          test_idx <- which(sy$type == test_dset_nm & sy$therapy_type == test_type)
          
          # 4. Evaluate metrics
          if (length(test_idx) > 0) {
            cat("Training:", train_dset_nm, ", Therapy:", train_type, " | Testing:", test_dset_nm, ", Therapy:", test_type, "\n")
            tmp <- evaluate_test_set(fit = enet_mdl, test_idx = test_idx, sy = sy, th_filt_names = feature_names, B = B)
            Op <- rbind(Op, data.frame(
              Train         = train_dset_nm,
              Test          = test_dset_nm,
              bmed          = tmp[1],
              lci           = tmp[2],
              uci           = tmp[3],
              therapy_test  = test_type,
              therapy_train = train_type
            ))    
          } else {
            cat("No samples of therapy", test_type, "in", test_dset_nm, "\n")
          }
          
          
        }
      }
      
    } else {
      cat("Not enough samples of therapy", train_type, "in", train_dset_nm, "\n")
      next
    }
    
  }
  
}

```

### Visualise
```{r}
study_order = c("GYN", "UGB", "NEN", "McCulloch_2022", "Lee_2022", "Matson_2018", "Frankel_2017")
Op$plot_value <- Op$bmed
Op$plot_value[Op$bmed == 0] <- NA

Op2 <- Op %>%
  filter(Train %in% study_order,
         Test %in% study_order) %>%
  group_by(therapy_train, therapy_test, Train) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup() %>%
  group_by(therapy_train, therapy_test, Test) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup()

Op2$Train <- factor(Op2$Train, levels = rev(study_order))  # top → bottom
Op2$Test  <- factor(Op2$Test,  levels = study_order)       # left → right

Op2$plot_value <- Op2$bmed

make_paper_heatmap(Op2)
```

## Load the metadata for rare cancers

```{r}
# Gunjur et al. data
meta_path_ash <- "./data/ash_pancancer/metadata_full.tsv"
meta_ash <- read.csv(meta_path_ash, sep = '\t') 
meta_ash = cbind(run_acc = meta_ash$run_accession, meta_ash)

# Outcome
meta_ash$RvsP = "R"
meta_ash$RvsP[which(meta_ash$BOR == "PD" | meta_ash$BOR == "cPD")] = "NR" 
meta_ash$RvsP = factor(meta_ash$RvsP, levels = c("NR", "R"))
```

### Load Sylph outputs

```{r}
sy_ash <- read_sylph("./data/ash_pancancer/combined_q_99.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy_ash) <- gsub("_1", "", colnames(sy_ash))
colData(sy_ash)$Sample_file <- gsub("_1", "", basename(colData(sy_ash)$Sample_file))

# Reorder meta_ash 
meta_ash = meta_ash[match(colnames(sy_ash), meta_ash$run_accession), ]
# get rid of SD
rmidx = which(meta_ash$BOR == "SD")
if(length(rmidx) > 0){
  meta_ash = meta_ash[-rmidx, ]
  sy_ash = sy_ash[, -rmidx]
}

sy_ash <- filter_by_presence(sy_ash, min_nonzero = 8) # filter at 10%
dim(sy_ash)

# Checks before merging metadata
all(colnames(sy_ash) %in% meta_ash$run_accession)
all(meta_ash$run_accession %in% colnames(sy_ash))

sy_ash = modify_metadata(sy_ash, meta_ash)

```

### Run StrainSpy

```{r}
design <- as.formula("~ RvsP + histology_cohort.x + age + sex + BMI") 

save_path <- "output_rds/ASH_zib_q_99_ebp.rds"

if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  # Run with weak prior for prediction
  ebp = compute_eb_priors(sy_ash, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
  ZB_fit <- glmZiBFit(sy_ash, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit, save_path)
}



# These are strainspy predictors from Pancx data
th = top_hits = top_hits(ZB_fit, coef = 2, alpha = 0.05)
## !!NB: We can't build a model using melanoma data and predict Ash's data - that will cause data leakage!!
# But we can build a model using Ash's data and predict Melanoma. 

# Since Ash's data is all CICB, we'll do cross testing on individual datasets by running strainspy separately and building models

```


### Load the Melanoma datasets

```{r}
meta_path <- "./data/ash_pancancer/metadata_full.tsv"
meta_pan <- read.csv(meta_path, sep = '\t') 
meta_pan = cbind(run_acc = meta_pan$run_accession, meta_pan)

# Outcome
meta_pan$RvsP = "R"
meta_pan$RvsP[which(meta_pan$BOR == "PD" | meta_pan$BOR == "cPD")] = "NR" 
meta_pan$RvsP = factor(meta_pan$RvsP, levels = c("NR", "R"))

# Melanoma meta - Thanks, Kshitij+Allyson (20260615)
meta_m = read.csv("data/melanoma_pooled/meta_melanoma_sm.csv")
# meta_m2 = read.csv("data/melanoma_pooled/Clean_metadata_Aug5_Allyson_match_2025.txt", sep = '\t')

##### Let's filter sequences to match with Fig. 4a #####
# see select_melanoma_samples.R
keep_samples = readLines("data/melanoma_pooled/gunjur_samples.txt")
meta_m_filt = meta_m %>% filter(X %in% keep_samples)

meta = rbind(data.frame(X = meta_pan$run_acc, c_type = "RARE", type = meta_pan$histology_cohort.x, therapy_type = "combination", RvsP = meta_pan$RvsP), 
             data.frame(X = meta_m_filt$X, c_type = "MEL", type = meta_m_filt$Study_simplified, therapy_type = meta_m_filt$ICB, RvsP = meta_m_filt$ORR))

meta$therapy_type <- ifelse(meta$therapy_type %in% c("aPD1/aCTLA", "combination"), "CICB", ifelse(meta$therapy_type == "aPD1", "PD1", "other"))
meta$therapy_type[is.na(meta$therapy_type)] = "other" # Get this of the NAs 


sy = read_sylph("./data/melanoma_pooled/combined_q_99.tsv") # q99
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

# Reduce sy to only keep the ones we have metadata for
sy = sy[, colnames(sy) %in% meta$X]

# Reorder meta 
meta = meta[match(colnames(sy), meta$X), ]

# Checks before merging metadata
all(colnames(sy) %in% meta$X)
all(meta$X %in% colnames(sy))

sy = modify_metadata(sy, meta, replace = T)

### Filter and keep Melanoma only
sy_melanoma = sy[,sy$c_type=="MEL"]
# sy_melanoma <- filter_by_presence(sy_melanoma, min_nonzero = 42) # filter at 10%
dim(sy_melanoma)

meta_melanoma = meta[meta$c_type == "MEL", ]

```

## Predict using StrainSpy significant predictors

### Build a model using rare cancer data

```{r}
th_filt <- th[th$Contig_name %in% rownames(sy_melanoma), ] # We are working with just 34 predictors

train_idx <- 1:dim(sy_ash)[2]
fit <- run_enet(train_idx, "", meta_ash, sy_ash, th_filt$Contig_name, return_fit = TRUE)

test_dset = unique(sy_melanoma$type)
therapy_types <- c("CICB", "PD1")

B <- 1000
set.seed(1988)

Op <- data.frame()

for (test_dset_nm in test_dset) {
  cat("Training:", "rare", "(", "CICB", ") | Testing set:", test_dset_nm, "\n")
  
  for (test_type in therapy_types) {
    # Determine test indices based on type criteria (preserving your original != 'CICB' logic for 'Other')
    test_idx <- which(sy_melanoma$type == test_dset_nm & sy_melanoma$therapy_type == test_type)
    
    # 4. Evaluate metrics
    if (length(test_idx) > 0) {
      tmp <- evaluate_test_set(fit, sy = sy, test_idx, th_filt$Contig_name, B = B)
      
    } else {
      tmp <- c(NA, NA, NA)
    }
    
    # 5. Append cleanly to output data frame
    Op <- rbind(Op, data.frame(
      Train         = 'rare', 
      Test          = test_dset_nm,
      bmed          = tmp[1], 
      lci           = tmp[2], 
      uci           = tmp[3], 
      therapy_test  = test_type, 
      therapy_train = 'CICB'
    ))
  }
}


ggplot(Op, aes(x = bmed, y = Test, color = therapy_test)) +
  # Adds a dashed reference line at AUC = 0.5 (random chance)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
  # Draws the CI error bars (dodged so overlapping test points don't clash)
  geom_errorbarh(aes(xmin = lci, xmax = uci), width = 0.3, 
                 position = position_dodge(width = 0.4), size = 0.8) +
  # Draws the median AUC points
  geom_point(position = position_dodge(width = 0.4), size = 3.5) +
  # Formatting & Labels
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Bootstrap Median AUC with 95% Confidence Intervals",
    subtitle = "Grouped by Testing Therapy Type",
    x = "Median AUC (bmed)",
    y = "Study (Test)",
    color = "Therapy Test"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold")
  )


```

### Training and testing

```{r, warning=FALSE, message=FALSE}
# Test and train on all melanoma
train_dset = unique(sy_melanoma$type)
test_dset = train_dset

Op <- data.frame()
th_filt <- th[th$Contig_name %in% rownames(sy_melanoma), ]
B <- 1000
set.seed(1988)

# Define the logical conditions to iterate through
therapy_types <- c("CICB", "PD1")

# --- Nested Iteration Loops ---
for (train_dset_nm in train_dset) {
  for (train_type in therapy_types) {
    
    # 1. Filter training indices dynamically
    train_idx <- which(sy_melanoma$type == train_dset_nm & sy_melanoma$therapy_type == train_type)
    
    # 2. Train model if data exists, otherwise prepare to log zeros
    has_train_data <- length(train_idx) > 0
    if (has_train_data) {
      fit <- run_enet(train_idx, "", meta_melanoma, sy_melanoma, th_filt$Contig_name, return_fit = TRUE)
    }
    
    # 3. Process test datasets
    for (test_dset_nm in test_dset) {
      cat("Training:", train_dset_nm, "(", train_type, ") | Testing set:", test_dset_nm, "\n")
      
      for (test_type in therapy_types) {
        # Determine test indices based on type criteria (preserving your original != 'CICB' logic for 'Other')
        test_idx <- which(sy_melanoma$type == test_dset_nm & sy_melanoma$therapy_type == test_type)
        # if (test_type == "combo") {
        #   test_idx <- which(sy_melanoma$type == test_dset_nm & sy_melanoma$therapy_type == "combo")
        # } else {
        #   test_idx <- which(sy_melanoma$type == test_dset_nm & sy_melanoma$therapy_type != "CICB")
        # }
        
        # 4. Evaluate metrics
        if (has_train_data && length(test_idx) > 0) {
          tmp <- evaluate_test_set(fit = fit, sy = sy_melanoma, test_idx = test_idx, th_filt_names = th_filt$Contig_name, B = B)
        } else {
          tmp <- c(NA, NA, NA)
        }
        
        # 5. Append cleanly to output data frame
        Op <- rbind(Op, data.frame(
          Train         = train_dset_nm, 
          Test          = test_dset_nm,
          bmed          = tmp[1], 
          lci           = tmp[2], 
          uci           = tmp[3], 
          therapy_test  = test_type, 
          therapy_train = train_type
        ))
      }
    }
  }
}
```

### Visualise

```{r, fig.width=12, fig.height=12, fig.align='center'}
# Try to mimic ash's original figure
# Unfortunately we can't access Simpson and Andrews, everything else is here (and a couple more as well!)
study_order <- c("McCulloch_2022", "Lee_2022", "Matson_2018", "Frankel_2017")

Op$plot_value <- Op$bmed
Op$plot_value[Op$bmed == 0] <- NA

Op2 <- Op %>%
  filter(Train %in% study_order,
         Test %in% study_order) %>%
  group_by(therapy_train, therapy_test, Train) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup() %>%
  group_by(therapy_train, therapy_test, Test) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup()

Op2$Train <- factor(Op2$Train, levels = rev(study_order))  # top → bottom
Op2$Test  <- factor(Op2$Test,  levels = study_order)       # left → right

Op2$plot_value <- Op2$bmed

make_paper_heatmap(Op2)

```

### Load data from Gunjur et al using AI

```{r}
# chatGPT pull from paper
Op4d <- data.frame(
  
  Train = c(
    # CICB → CICB
    rep("CA209-538", 9),
    rep("2022_Simpson", 9),
    rep("2022_Lee", 9),
    rep("2021_Andrews", 9),
    rep("2017_Frankel", 9),
    
    # PD1 → CICB + PD1
    rep("2022_McCulloch", 9),
    rep("2022_Lee_PD1", 9),
    rep("2018_Matson", 9),
    rep("2017_Frankel_PD1", 9)
  ),
  
  Test = c(
    rep(c("CA209-538","2022_Simpson","2022_Lee","2021_Andrews","2017_Frankel",
          "2022_McCulloch","2022_Lee_PD1","2018_Matson","2017_Frankel_PD1"), each = 1)
  ),
  
  bmed = c(
    
    # CICB train → CICB test (5 cols) + CICB→PD1 (4 cols)
    1, 0.67, 0.40, 0.78, 0.75,  0.46, 0.44, 0.58, 0.54,
    0.57, 1, 0.52, 0.59, 0.59,   0.50, 0.52, 0.42, 0.25,
    0.39, 0.64, 1, 0.43, 0.35,   0.36, 0.59, 0.23, 0.61,
    0.72, 0.65, 0.52, 1, 0.63,   0.40, 0.41, 0.46, 0.21,
    0.62, 0.74, 0.47, 0.65, 1,   0.40, 0.50, 0.40, 0.29,
    
    # PD1 train → CICB test (5 cols) + PD1 test (4 cols)
    0.53, 0.29, 0.26, 0.58, 0.55,  1, 0.52, 0.57, 0.64,
    0.42, 0.63, 0.66, 0.22, 0.21,  0.58, 1, 0.61, 0.71,
    0.42, 0.40, 0.41, 0.39, 0.42,  0.56, 0.48, 1, 0.64,
    0.35, 0.33, 0.51, 0.37, 0.30,  0.66, 0.49, 0.65, 1
  )
)

Op4d$therapy_train <- ifelse(grepl("McCulloch|Lee_PD1|Matson|Frankel_PD1", Op4d$Train),
                             "PD1", "CICB")

Op4d$therapy_test <- ifelse(grepl("McCulloch|Lee_PD1|Matson|Frankel_PD1", Op4d$Test),
                            "PD1", "CICB")


Op4d$Train[grep("McCulloch", Op4d$Train)] = "McCulloch_2022"
Op4d$Test[grep("McCulloch", Op4d$Test)] = "McCulloch_2022"

Op4d$Train[grep("Lee", Op4d$Train)] = "Lee_2022"
Op4d$Test[grep("Lee", Op4d$Test)] = "Lee_2022"

Op4d$Train[grep("Matson", Op4d$Train)] = "Matson_2018"
Op4d$Test[grep("Matson", Op4d$Test)] = "Matson_2018"

Op4d$Train[grep("Frankel", Op4d$Train)] = "Frankel_2017"
Op4d$Test[grep("Frankel", Op4d$Test)] = "Frankel_2017"

Op_ash = Op4d %>% filter(Train %in% study_order,
                         Test %in% study_order)

Op_ash$plot_value <- Op_ash$bmed
Op_ash$plot_value[Op_ash$bmed == 0] <- NA

Op_ash2 <- Op_ash %>%
  filter(Train %in% study_order,
         Test %in% study_order) %>%
  group_by(therapy_train, therapy_test, Train) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup() %>%
  group_by(therapy_train, therapy_test, Test) %>%
  filter(any(!is.na(plot_value))) %>%
  ungroup()

Op_ash2$Train <- factor(Op_ash2$Train, levels = rev(study_order))  # top → bottom
Op_ash2$Test  <- factor(Op_ash2$Test,  levels = study_order)       # left → right

Op_ash2$plot_value <- Op_ash2$bmed

# Original-ish
make_paper_heatmap(Op_ash2)

```

### Build a comparison plot of better visualisation

```{r}

df_new <- Op2 %>%
  transmute(
    Train,
    Test,
    therapy_train,
    therapy_test,
    cANI = bmed
  )

df_paper <- Op_ash2 %>%
  transmute(
    Train,
    Test,
    therapy_train,
    therapy_test,
    Paper = bmed
  )

df_pair <- inner_join(df_new, df_paper,
                      by = c("Train", "Test", "therapy_train", "therapy_test")) %>%
  filter(Train != Test) %>% # diagonal is annoying and skews everything 
  mutate(therapy_pair = paste(therapy_train, ">", therapy_test))

ggplot(df_pair, aes(x = Paper, y = cANI)) +
  # 1. Move the identity line to the background layer so it doesn't cut through points
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60", alpha = 0.8) +
  
  # 2. Consolidated into a single point layer with clean editorial colors
  geom_point(aes(fill = cANI > Paper), shape = 21, color = "black", size = 2.5, stroke = 0.6, alpha = 0.9) +
  
  # 3. Use ggrepel to dynamically push text away from points and other text blocks
  geom_text_repel(
    aes(label = paste0(Train, " → ", Test)),
    size = 3.2,
    box.padding = 0.3,
    point.padding = 0.3,
    segment.color = "gray70",
    segment.size = 0.4,
    min.segment.length = 0, # Always draw lines to points if they move
    max.overlaps = 15       # Increase or set to Inf if you absolutely want every label shown
  ) +
  
  # 4. Custom color scale (Clear visual cue: Blue = StrainSpy won, Orange/Red = Baseline won)
  scale_fill_manual(
    values = c("TRUE" = "#2b8cbe", "FALSE" = "#fdbb84"),
    labels = c("TRUE" = "StrainSpy + cANI", "FALSE" = "Paper Baseline"),
    name = "Higher AUC Model:"
  ) +
  
  # 5. Fix coordinates and facet panels
  coord_equal() +
  facet_grid(therapy_train ~ therapy_test) +
  
  # 6. Use theme_bw for clean panel borders, then finely tune the typography
  theme_bw(base_size = 14) + 
  labs(
    y = "StrainSpy + cANI (AUC)",
    x = "All Genomes + Abundance (AUC)",
    title = "AUC Comparison Across All Train–Test Settings",
    subtitle = "Points above the dashed line indicate StrainSpy outperforming the baseline paper"
  ) +
  theme(
    # Typography
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 11, color = "gray40", margin = margin(b = 12)),
    axis.title = element_text(face = "bold", size = 12),
    
    # Clean up Facet Header Boxes (Strips)
    strip.background = element_rect(fill = "gray95", color = "gray80"),
    strip.text = element_text(face = "bold", size = 11, color = "gray20"),
    
    # Grid adjustments
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.2, "lines"), # Give facets room to breathe so text doesn't bleed across boxes
    
    # Legend Placement
    legend.position = "bottom",
    legend.box.spacing = unit(0.5, "cm"),
    legend.text = element_text(size = 11)
  )
```

## Prediction using LODO


# StrainSpy on the Pan-cancer model (rare + Melanoma)

## Load metadata 

```{r}
meta_path <- "./data/ash_pancancer/metadata_full.tsv"
meta_pan <- read.csv(meta_path, sep = '\t') 
meta_pan = cbind(run_acc = meta_pan$run_accession, meta_pan)
# From paper:
# RvsP = CR or PR VS. PD or cPD - excluded patients with a BOR of stable disease (SD) (n=29)

# Outcome
meta_pan$RvsP = "R"
meta_pan$RvsP[which(meta_pan$BOR == "PD" | meta_pan$BOR == "cPD")] = "NR" 
meta_pan$RvsP = factor(meta_pan$RvsP, levels = c("NR", "R"))

# Melanoma meta
meta_m = read.csv("data/melanoma_pooled/Clean_metadata_Aug5_Allyson_match_2025.txt", sep = '\t')


meta = rbind(data.frame(X = meta_pan$run_acc, c_type = "RARE", type = meta_pan$histology_cohort.x, RvsP = meta_pan$RvsP), 
             data.frame(X = meta_m$X, c_type = "MEL", type = meta_m$Study_simplified, RvsP = meta_m$ORR))

```

## Load sylph outputs

```{r}
sy <- read_sylph("./data/ash_pancancer/merged_pan_melanoma.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

# Reorder meta 
meta = meta[match(colnames(sy), meta$X), ]

sy <- filter_by_presence(sy, min_nonzero = 53) # filter at 10%
dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$X)
all(meta$X %in% colnames(sy))

sy = modify_metadata(sy, meta, replace = T)
```

## Fit the univariate model - combined cancers

```{r}
design <- as.formula("~ RvsP")

save_path <- "output_rds/ASH+mela_zib_q_99_ebp.rds"

if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  # Run with weak prior for prediction
  ebp = compute_eb_priors(sy, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
  ZB_fit <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit, save_path)
}

top_hits(ZB_fit) # No hits - Melanoma is quite heterogenous
```

## Build a Predictive Species list

### Plot this on a tree

```{r, fig.width=12, fig.height=12}
th = top_hits(ZB_fit, alpha = 1)
th$min_p <- pmin(th$p_value, th$zi_p_value)
th <- th[order(th$min_p), ]
tax_99 = read_taxonomy("data/TAXONOMY/sylph_DB_taxonomy_99.tsv")
th2 = strainspy:::add_tax2tophits(th, tax_99, columns = c("Phylum","Genus","Species"))

# Fit a model using all data and take coefficients
f = 500
fit = run_enet(train_idx = 1:dim(sy)[2], test_idx = "", return_fit = T, meta = meta, sy = sy, feature_names = th$Contig_name[1:f])
coefs <- coef(fit$finalModel, s = fit$bestTune$lambda); coefs = coefs[coefs[,1] != 0,]; coefs = coefs[-1]

# write.table(names(coefs), quote = F, row.names = F, col.names = F, "pancancer_tree_accessions.txt") 

# tree

tree = read.tree("data/ash_pancancer/tree.nwk")
tree$tip.label = unname(sapply(tree$tip.label, function(x) paste(strsplit(x, "_")[[1]][1:2], collapse = "_")))
tree = phytools::midpoint.root(tree)

df <- th2 %>%
  filter(Genome_file %in% names(coefs)) %>%
  transmute(
    tip     = Genome_file,
    Phylum  = Phylum,
    Genus   = Genus,
    Species = Species,
    Coef    = coefs[Genome_file]   # attach coefficients directly
  ) %>%
  arrange(match(tip, tree$tip.label))  # order to match tree

df = as.data.frame(df)

rownames(df) = df$tip

com_gen <- c(
  "Ruminococcus_B",
  "Faecalibacterium",
  "Phascolarctobacterium",
  "Bacteroides",
  "Actinomyces",
  "Alistipes",
  "Gemmiger",
  "Clostridium_A",
  "Collinsella",
  "Veillonella",
  "Streptococcus"
)

genus_cols <- c(
  "#000000", "#f781bf", "#ff7f00",
  "#009E73", "#984ea3", "#4daf4a",
  "#ffff33", "#a65628", "#005999", 
  "#377eb8", "#E41A1C", "#DDDDDD"
)

names(genus_cols) = c(com_gen, "Other")

df$Genus[!df$Genus %in% com_gen] = "Other"

df$Genus = factor(df$Genus, levels = c(com_gen, "Other"))


df = df[rownames(df) %in% tree$tip.label,]
tree = ape::keep.tip(tree,df$tip)

p <- ggtree(tree, layout = "circular")
p$data <- merge(
  p$data,
  df,
  by.x = "label",
  by.y = "tip",
  all.x = TRUE
)

p <- p +
  geom_tippoint(
    aes(color = Genus),
    size = 2
  )  +
  scale_color_manual(
    values = genus_cols,
    name = "Genus"
  )




df_tmp <- data.frame(Coef = df$Coef)
rownames(df_tmp) <- df$tip

p <- gheatmap(
  p,
  df_tmp,
  offset = 0,
  width = 0.05,
  colnames = FALSE
) +
  scale_fill_gradient2(
    high = "#084594",
    mid = "white",
    low = "#99000D",
    midpoint = 0,
    transform = "pseudo_log",
    name = "ENet coef"
  ) +
  theme(
    text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p

```

### Scatter plot with line to see the variation

```{r, fig.width=12}
df2 = df
med_df <- df2 %>%
  filter(Genus != "Other") %>%
  group_by(Genus) %>%
  summarise(
    med_coef = median(Coef, na.rm = TRUE),
    .groups = "drop"
  )


scfa_gen <- c(
  # beneficial / SCFA
  "Faecalibacterium",
  "Ruminococcus_B",
  "Phascolarctobacterium",
  "Gemmiger",
  "Eubacterium_F",
  "Bifidobacterium")

genus_order <- df2 %>%
  filter(Genus != "Other") %>%
  group_by(Genus) %>%
  summarise(
    med_coef = median(Coef, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    is_scfa = Genus %in% scfa_gen
  ) %>%
  arrange(
    desc(is_scfa),   # SCFA first
    desc(med_coef)  # then strictly decreasing coef
  ) %>%
  pull(Genus)



med_df$Genus <- factor(med_df$Genus, levels = genus_order)
df2$Genus <- factor(df2$Genus, levels = med_df$Genus[order(med_df$med_coef, decreasing = T)])

fit <- lm(med_coef ~ as.numeric(factor(Genus, levels = med_df$Genus)), data = med_df)

# Coefficients
coef(fit)

ggplot(df2, aes(x = Genus, y = Coef, color = Genus, colour = genus_cols)) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.25
  ) +
  geom_jitter(
    width = 0.15,
    size = 2,
    alpha = 0.8
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_smooth(
    data = med_df,
    aes(y = med_coef, group = 1),
    method = "lm",
    se = FALSE,
    colour = "black",
    linewidth = 0.9
  ) +
  scale_color_manual(
    values = genus_cols,
    guide = "none"   # legend already in the tree
  ) +
  labs(
    x = "Ordered Genus",
    y = "Elastic-net coefficient"
  ) +
  ggthemes::theme_clean(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
