---
title: "Associations between gut microbial strains and cancer immunotherapy outcomes - a study in rare cancers and melanoma"
output:
  github_document
date: "2026-06-20"
editor_options:
  chunk_output_type: console
---
## Load dependencies

```{r, warning=FALSE, message=FALSE}
library(strainspy)
library(SummarizedExperiment)
library(caret)
library(glmnet) # elastic net
library(ggrepel)
library(ranger) # RF
library(pROC)
library(doParallel)
library(foreach)
library(parallel)
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggtree)
library(ggsci)
```

# Association testing

Classic StrainSpy applied to rare cancer and melanoma data in different combinations

## Load the metadata for rare cancers

```{r}
# Gunjur et al. data
meta_path_ash <- "./data/ash_pancancer/metadata_full.tsv"
meta_ash <- read.csv(meta_path_ash, sep = '\t') 
meta_ash = cbind(run_acc = meta_ash$run_accession, meta_ash)

# Outcome
meta_ash$RvsP = "R"
meta_ash$RvsP[which(meta_ash$BOR == "PD" | meta_ash$BOR == "cPD")] = "NR" 
meta_ash$RvsP = factor(meta_ash$RvsP, levels = c("NR", "R"))
```

### Load Sylph outputs

```{r}
sy_ash <- read_sylph("./data/ash_pancancer/combined_q_99.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy_ash) <- gsub("_1", "", colnames(sy_ash))
colData(sy_ash)$Sample_file <- gsub("_1", "", basename(colData(sy_ash)$Sample_file))

# Reorder meta_ash 
meta_ash = meta_ash[match(colnames(sy_ash), meta_ash$run_accession), ]
# get rid of SD
rmidx = which(meta_ash$BOR == "SD")
if(length(rmidx) > 0){
  meta_ash = meta_ash[-rmidx, ]
  sy_ash = sy_ash[, -rmidx]
}

sy_ash <- filter_by_presence(sy_ash, min_nonzero = 8) # filter at 10%
dim(sy_ash)

# Checks before merging metadata
all(colnames(sy_ash) %in% meta_ash$run_accession)
all(meta_ash$run_accession %in% colnames(sy_ash))

sy_ash = modify_metadata(sy_ash, meta_ash)

```

### Run StrainSpy

```{r}
design <- as.formula("~ RvsP + histology_cohort.x + age + sex + BMI") 

save_path <- "output_rds/ASH_zib_q_99_ebp.rds"

if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  # Run with weak prior for prediction
  ebp = compute_eb_priors(sy_ash, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
  ZB_fit <- glmZiBFit(sy_ash, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit, save_path)
}

# These are strainspy predictors from Pancx data
th = top_hits = top_hits(ZB_fit, coef = 2, alpha = 0.05)

```

## Inspect StrainSpy hits species/genera
```{r}
tax_99 = read_taxonomy("data/TAXONOMY/sylph_DB_taxonomy_99.tsv")

th_ph = comp_ani_diff_and_posthoc_test(sy_ash, ZB_fit, th)

# Interestingly, these are all beta hits (differences in strain identity)
table(th_ph$hit_component)

th_ph_full = strainspy:::add_tax2tophits(th_ph, tax_99, c("Genus", "Species"))

tbl_full = table(th_ph_full$Species)
sort(tbl_full[tbl_full>5], decreasing = T)

tbl_full = table(th_ph_full$Genus)
sort(tbl_full[tbl_full>5], decreasing = T)

# Although there are interesting looking strains, we should remove poorly supported ones
th_ph_filt = th_ph_full[is.na(th_ph_full$Comment), ]

tbl = table(th_ph_filt$Species)
sort(tbl, decreasing = T)

tbl = table(th_ph_filt$Genus)
sort(tbl, decreasing = T)
```

# Inspect effect sizes

```{r, fig.width=12}
th_ph_ord = th_ph_filt[order(abs(th_ph_filt$ANI_Difference) , decreasing = T),]

summary_tbl <- th_ph_filt %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(
    N_hits = n(),
    
    Top_adj_p_beta  = min(p_adjust, na.rm = TRUE),
    Top_coef_beta   = coefficient[which.min(p_adjust)],
    Top_se_beta     = std_error[which.min(p_adjust)],
    Top_contig_beta = Contig_name[which.min(p_adjust)],
    
    Top_adj_p_ZI    = min(zi_p_adjust, na.rm = TRUE),
    Top_coef_ZI     = zi_coefficient[which.min(zi_p_adjust)],
    Top_se_ZI       = zi_std_error[which.min(zi_p_adjust)],
    Top_contig_ZI   = Contig_name[which.min(zi_p_adjust)],
    
    
    .groups = "drop"
  ) %>%
  mutate(
    Min_adj_p = pmin(Top_adj_p_beta, Top_adj_p_ZI, na.rm = TRUE),
    Dominant_component = ifelse(Top_adj_p_beta < Top_adj_p_ZI, "Beta", "ZI"),
    Dominant_contig    = ifelse(Dominant_component == "Beta", Top_contig_beta, Top_contig_ZI),
    Dominant_coef      = ifelse(Dominant_component == "Beta", Top_coef_beta, Top_coef_ZI),
    Dominant_se        = ifelse(Dominant_component == "Beta", Top_se_beta, Top_se_ZI)
  ) %>%
  select(Species, N_hits, Dominant_component, Min_adj_p, Dominant_coef, Dominant_se, Dominant_contig) %>%
  arrange(Min_adj_p)

beta_summary <- summary_tbl %>%
  filter(Dominant_component == "Beta") %>%
  arrange(Min_adj_p) %>%
  mutate(
    CI_low = Dominant_coef - 1.96 * Dominant_se,
    CI_high = Dominant_coef + 1.96 * Dominant_se,
    Trend = ifelse(Dominant_coef > 0, "\u2191 R (similar)", "\u2193 R (divergent)"),
    Species = factor(Species, levels = rev(unique(Species)))
  ) %>%
  arrange(desc(Dominant_coef)) %>%
  mutate(Species = factor(Species, levels = rev(unique(Species))))

ggplot(beta_summary, aes(x = Dominant_coef, y = Species, color = Trend)) +
  geom_point(aes(size = N_hits)) +
  geom_errorbar(aes(xmin = CI_low, xmax = CI_high), width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("\u2191 R (similar)" = "red",
                                "\u2193 R (divergent)" = "blue")) +
  scale_size_continuous(
    name = "Strains",
    range = c(2, 8),
    breaks = c(1, 5, 20),
    labels = c("1", "5", "20")
  ) +
  labs(
    x = "Effect size (difference in ANI, NR vs R)",
    y = "Species",
    color = "Trend"
  ) +
  theme_minimal(base_size = 16)

# Some of these hits appear interesting
plot_ani_dist(sy_ash, 'RvsP', beta_summary$Dominant_contig[1:7])
plot_ani_dist(sy_ash, 'RvsP', beta_summary$Dominant_contig[8:14])


```

Hits in Rothia sp, specifically Rothia mucilaginosa, previously detected modulating CRC immunotherapy response - reported previously is an abundance (presence/absence) signal, but we see a difference in strain identity - https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0307639

Hits in Phascolarctobacterium and Ruminococcus have been previously detected modulating lung cancer immunotherapy response - https://link.springer.com/article/10.1186/s40164-023-00442-x

Bacteroides sp. - well studied and understood role in modulating CTLA-4 blockade - https://www.science.org/doi/10.1126/science.aad1329 - our hit is from a weird and less known species - but the genus is correct. ANI can theoretically tag links like this.

There are some other papers discussing Phascolarctobacterium and Clostridium hits, but they don’t look very specific and our hits are again from not well classified species (https://www.nature.com/articles/s41586-019-0878-z, https://www.nature.com/articles/s41591-022-01694-6, https://www.tandfonline.com/doi/full/10.1080/2162402X.2022.2081010)

## Rare cancer + Melanoma - combined dataset

```{r}
meta_path <- "./data/ash_pancancer/metadata_full.tsv"
meta_pan <- read.csv(meta_path, sep = '\t') 
meta_pan = cbind(run_acc = meta_pan$run_accession, meta_pan)
# From paper:
# RvsP = CR or PR VS. PD or cPD - excluded patients with a BOR of stable disease (SD) (n=29)

# Outcome
meta_pan$RvsP = "R"
meta_pan$RvsP[which(meta_pan$BOR == "PD" | meta_pan$BOR == "cPD")] = "NR" 
meta_pan$RvsP = factor(meta_pan$RvsP, levels = c("NR", "R"))

# Melanoma meta
meta_m = read.csv("data/melanoma_pooled/meta_melanoma_sm.csv")

meta = rbind(data.frame(X = meta_pan$run_acc, c_type = "RARE", type = meta_pan$histology_cohort.x, therapy_type = "combination", RvsP = meta_pan$RvsP), 
             data.frame(X = meta_m$X, c_type = "MEL", type = meta_m$Study_simplified, therapy_type = meta_m$ICB, RvsP = meta_m$ORR))
meta$therapy_type <- ifelse(meta$therapy_type %in% c("aPD1/aCTLA", "combination"), "CICB", ifelse(meta$therapy_type == "aPD1", "PD1", "other"))
meta$therapy_type[is.na(meta$therapy_type)] = "other" # Get this of the NAs 

sy <- read_sylph("./data/ash_pancancer/merged_pan_melanoma.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

# Reorder meta 
meta = meta[match(colnames(sy), meta$X), ]

sy <- filter_by_presence(sy, min_nonzero = 53) # filter at 10%
dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$X)
all(meta$X %in% colnames(sy))

sy = modify_metadata(sy, meta, replace = T)

```

### Other association tests 

```{r}
#### DO NOT RE-RUN BELOW - NO HITS ####


## Full dataset (526 samples) ### No hits for this ##
# design <- as.formula("~ RvsP") # no hits
# design <- as.formula("~ RvsP + c_type") # no hits
# design <- as.formula("~ RvsP + type") # no hits
# design <- as.formula("~ RvsP + type + therapy_type") # no hits
# 
# if(file.exists(save_path)){
#   ZB_fit <- readRDS(save_path)
# } else {
#   # Run with weak prior for prediction
#   ebp = compute_eb_priors(sy, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
#   ZB_fit <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
#   saveRDS(ZB_fit, save_path)
# }


# Drop Spencer and retry (359 samples)
# sy_nospecer = sy[,sy$type!="Spencer_2021"]
# sy_nospecer <- filter_by_presence(sy_nospecer, min_nonzero = 36) # filter at 10%
# 
# # Drop rare cancers + spencer and retry
# sy_nospecer_norare = sy_nospecer[,sy_nospecer$c_type!="RARE"]
# sy_nospecer_norare <- filter_by_presence(sy_nospecer_norare, min_nonzero = 25) # filter at 10%
# 
# design <- as.formula("~ RvsP + type + therapy_type") # still no hits
# 
# save_path <- "output_rds/ASH+mela_zib_q_99_ebp.rds"
# 
# if(file.exists(save_path)){
#   ZB_fit <- readRDS(save_path)
# } else {
#   # Run with weak prior for prediction
#   ebp = compute_eb_priors(sy_nospecer_norare, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
#   ZB_fit <- glmZiBFit(sy_nospecer_norare, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
#   saveRDS(ZB_fit, save_path)
# }

# Try individual melanoma datasets as well - There are some hits here
op_th_melanoma_ind = data.frame()
sets = c("Frankel_2017","Gopalakrishnan_2018","Lee_2022","Matson_2018","McCulloch_2022","Spencer_2021")
design <- as.formula("~ RvsP") 
for (set in sets){
  sy1 = sy[,which(sy$type == set)]
  dim(sy1)
  sy1 = filter_by_presence(sy1, ceiling(dim(sy1)[2]/10))
  
  save_path <- file.path("output_rds", paste("melanoma_", set, ".rds", sep = ""))
  # design <- as.formula("~ RvsP") # no hits
  if(file.exists(save_path)){
    ZB_fit <- readRDS(save_path)
  } else {
    # Run with weak prior for prediction
    ebp = compute_eb_priors(sy1, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
    ZB_fit <- glmZiBFit(sy1, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
    saveRDS(ZB_fit, save_path)
  }
  th_out = top_hits(ZB_fit)
  if(nrow(th_out) > 0)
  {
    th_out = comp_ani_diff_and_posthoc_test(sy1, ZB_fit, th_out) # drop poorly supported ones on the fly
    op_th_melanoma_ind = rbind(op_th_melanoma_ind, 
                               cbind(th_out, dset = set)
    )
  }
}

op_th_melanoma_ind = op_th_melanoma_ind[is.na(op_th_melanoma_ind$Comment), ]
op_th_melanoma_ind = strainspy:::add_tax2tophits(op_th_melanoma_ind, taxonomy = tax_99, c("Genus", "Species"))
``` 

Looks like there are some hits in (1) Frankel_2017 (2) Gopalakrishnan_2018 (3) Matson_2018 

### Visualise

```{r,fig.width=12}

plot_ani_dist(sy[,which(sy$type == "Frankel_2017")], phenotype = 'RvsP', op_th_melanoma_ind$Contig_name[op_th_melanoma_ind$dset == "Frankel_2017"])
plot_ani_dist(sy[,which(sy$type == "Gopalakrishnan_2018")], phenotype = 'RvsP', op_th_melanoma_ind$Contig_name[op_th_melanoma_ind$dset == "Gopalakrishnan_2018"])
plot_ani_dist(sy[,which(sy$type == "Matson_2018")], phenotype = 'RvsP', op_th_melanoma_ind$Contig_name[op_th_melanoma_ind$dset == "Matson_2018"])

```

They look a bit different. Retest with therapy type as a covariate

```{r, fig.width=12}

op_th_melanoma_ind_tt = data.frame()
sets = c("Frankel_2017","Gopalakrishnan_2018","Lee_2022","Matson_2018","McCulloch_2022","Spencer_2021")
design <- as.formula("~ RvsP + therapy_type") 
for (set in sets){
  sy1 = sy[,which(sy$type == set)]
  dim(sy1)
  sy1 = filter_by_presence(sy1, ceiling(dim(sy1)[2]/10))
  
  ttbl = table(sy1$therapy_type)
  
  if(min(c( ttbl[['CICB']], ttbl[['PD1']]))  > 0.1*dim(sy1)[2]){ # 10% minimum rule
    save_path <- file.path("output_rds", paste("melanoma_tt_", set, ".rds", sep = ""))
    # design <- as.formula("~ RvsP") # no hits
    if(file.exists(save_path)){
      ZB_fit <- readRDS(save_path)
    } else {
      # Run with weak prior for prediction
      ebp = compute_eb_priors(sy1, design, nthreads = 10L, low_cutoff = 0, high_cutoff = Inf)
      ZB_fit <- glmZiBFit(sy1, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
      saveRDS(ZB_fit, save_path)
    }
    th_out = top_hits(ZB_fit)
    if(nrow(th_out) > 0)
    {
      th_out = comp_ani_diff_and_posthoc_test(sy1, ZB_fit, th_out) # drop poorly supported ones on the fly
      op_th_melanoma_ind_tt = rbind(op_th_melanoma_ind_tt, 
                                    cbind(th_out, dset = set)
      )
    }
  } else {
    cat("Not enough samples for the two therapy types")
  }
  
}

op_th_melanoma_ind_tt = op_th_melanoma_ind_tt[is.na(op_th_melanoma_ind_tt$Comment), ]

op_th_melanoma_ind_tt = strainspy:::add_tax2tophits(op_th_melanoma_ind_tt, taxonomy = tax_99, c("Genus", "Species"))

# Only hits in Frankel
plot_ani_dist(sy[,which(sy$type == "Frankel_2017")], phenotype = 'RvsP', op_th_melanoma_ind_tt$Contig_name[op_th_melanoma_ind_tt$dset == "Frankel_2017"])

```

### Summary plots

```{r, fig.width=12}

summary_tbl_mel <- op_th_melanoma_ind %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(
    N_hits = n(),
    
    Top_adj_p_beta  = min(p_adjust, na.rm = TRUE),
    Top_coef_beta   = coefficient[which.min(p_adjust)],
    Top_se_beta     = std_error[which.min(p_adjust)],
    Top_contig_beta = Contig_name[which.min(p_adjust)],
    
    Top_adj_p_ZI    = min(zi_p_adjust, na.rm = TRUE),
    Top_coef_ZI     = zi_coefficient[which.min(zi_p_adjust)],
    Top_se_ZI       = zi_std_error[which.min(zi_p_adjust)],
    Top_contig_ZI   = Contig_name[which.min(zi_p_adjust)],
    
    
    .groups = "drop"
  ) %>%
  mutate(
    Min_adj_p = pmin(Top_adj_p_beta, Top_adj_p_ZI, na.rm = TRUE),
    Dominant_component = ifelse(Top_adj_p_beta < Top_adj_p_ZI, "Beta", "ZI"),
    Dominant_contig    = ifelse(Dominant_component == "Beta", Top_contig_beta, Top_contig_ZI),
    Dominant_coef      = ifelse(Dominant_component == "Beta", Top_coef_beta, Top_coef_ZI),
    Dominant_se        = ifelse(Dominant_component == "Beta", Top_se_beta, Top_se_ZI)
  ) %>%
  select(Species, N_hits, Dominant_component, Min_adj_p, Dominant_coef, Dominant_se, Dominant_contig) %>%
  arrange(Min_adj_p)

beta_summary_mel <- summary_tbl_mel %>%
  filter(Dominant_component == "Beta") %>%
  arrange(Min_adj_p) %>%
  mutate(
    CI_low = Dominant_coef - 1.96 * Dominant_se,
    CI_high = Dominant_coef + 1.96 * Dominant_se,
    Trend = ifelse(Dominant_coef > 0, "\u2191 R (similar)", "\u2193 R (divergent)"),
    Species = factor(Species, levels = rev(unique(Species)))
  ) %>%
  arrange(desc(Dominant_coef)) %>%
  mutate(Species = factor(Species, levels = rev(unique(Species))))

ggplot(beta_summary_mel, aes(x = Dominant_coef, y = Species, color = Trend)) +
  geom_point(aes(size = N_hits)) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("\u2191 R (similar)" = "red",
                                "\u2193 R (divergent)" = "blue")) +
  scale_size_continuous(
    name = "Strains",
    range = c(2, 8),
    breaks = c(1, 5, 20),
    labels = c("1", "5", "20")
  ) +
  labs(
    x = "Effect size (difference in ANI, NR vs R)",
    y = "Species",
    color = "Trend"
  ) +
  theme_minimal(base_size = 16)

```