---
title: "Re-analysing pancancer dataset with dispersion priors"
output:
  github_document
date: "2026-04-27"
editor_options:
  chunk_output_type: console
---
## Load dependencies

```{r, warning=FALSE, message=FALSE}
library(strainspy)
library(SummarizedExperiment)
library(caret)
library(glmnet) # elastic net
library(ranger) # RF
library(pROC)
library(doParallel)
library(foreach)
library(parallel)
library(ggplot2)
library(tidyr)
library(dplyr)

```

## Load metadata 

```{r}
# WARNING! THIS INCLUDES UNPIBLISHED DATA
meta_path <- "./data/ash_pancancer/metadata_full.tsv"
meta <- read.csv(meta_path, sep = '\t') 
meta = cbind(run_acc = meta$run_accession, meta)
# From paper:
# RvsP = CR or PR VS. PD or cPD - excluded patients with a BOR of stable disease (SD) (n=29)

# Outcome
meta$RvsP = "R"
meta$RvsP[which(meta$BOR == "PD" | meta$BOR == "cPD")] = "NR" 
meta$RvsP = factor(meta$RvsP, levels = c("NR", "R"))
```

## Load sylph outputs

```{r}
sy <- read_sylph("./data/ash_pancancer/combined_q_99.tsv.gz") # q99)
# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

# Reorder meta 
meta = meta[match(colnames(sy), meta$run_accession), ]
# get rid of SD
rmidx = which(meta$BOR == "SD")
if(length(rmidx) > 0){
  meta = meta[-rmidx, ]
  sy = sy[, -rmidx]
}

sy <- filter_by_presence(sy, min_nonzero = 8) # filter at 10%
dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$run_accession)
all(meta$run_accession %in% colnames(sy))

sy = modify_metadata(sy, meta)
```

## Empirical Bayes priors without dispersion term

```{r}
design <- as.formula("~ RvsP + histology_cohort.x + age + sex + BMI") 

save_path <- "output_rds/ASH_prior_analysis_zib_q_99_ebp.rds"

if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy, design, nthreads = parallel::detectCores(), low_cutoff = 0,
                          high_cutoff = Inf)
  # Run with weak prior for prediction
  ZB_fit <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit, save_path)
}
```

There are 167 top hits, some of them dodgy
```{r}
th = top_hits(ZB_fit) 

# Looking at the first 10
plot_ani_dist(sy, 'RvsP', th$Contig_name[1:10]) 

```

Clearly, at least *B. animalis* and *F. nucleatum* should NOT be picked...

### Post-hoc testing

```{r}
th_ph = strainspy::comp_ani_diff_and_posthoc_test(sy, ZB_fit, th)

table(th_ph$Comment, useNA = 'always') # now there are only 16 hits

plot_ani_dist(sy, 'RvsP', th_ph$Contig_name[is.na(th_ph$Comment)]) 
```

## Empirical Bayes prior with dispersion terms

### Weak
```{r}
save_path <- "output_rds/ASH_prior_analysis_zib_q_99_ebp_disp5.rds"
ebp = ZB_fit@priors # priors are the same

# manually add the dispersion parameter
ebp@priors_df = rbind(ebp@priors_df, data.frame(prior = 'normal(0,5)', class = 'fixef_disp', coef = 1))

ebp@priors_df

if(file.exists(save_path)){
  ZB_fit_d5 <- readRDS(save_path)
} else {
  ZB_fit_d5 <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit_d5, save_path)
}

th_d5 = top_hits(ZB_fit_d5)

```

We still have 150 top hits, prior is likely too weak. Sensitivity analysis:

```{r}
plot_ani_dist(sy, 'RvsP', th_d5$Contig_name[1:20]) 
```

*B. animalis* is also still there...


### Strong
```{r}
save_path <- "output_rds/ASH_prior_analysis_zib_q_99_ebp_disp1.rds"
ebp = ZB_fit@priors # priors are the same

# manually add the dispersion parameter
ebp@priors_df = rbind(ebp@priors_df, data.frame(prior = 'normal(0,1)', class = 'fixef_disp', coef = 1))

ebp@priors_df

if(file.exists(save_path)){
  ZB_fit_d1 <- readRDS(save_path)
} else {
  ZB_fit_d1 <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit_d1, save_path)
}

th_d1 = top_hits(ZB_fit_d1)

```

Now there are no hits. This has killed all signal. Let's try with an empirical approach.

### Ebp prior

```{r}
save_path <- "output_rds/ASH_prior_analysis_zib_q_99_ebp_disp_ebp.rds"
ebp =  compute_eb_priors(sy, design, nthreads = parallel::detectCores(), low_cutoff = 0,
                          high_cutoff = Inf, est_disperion_prior = T)

plot_prior_bootstrap(ebp, 'dispersion')
ebp@priors_df

if(file.exists(save_path)){
  ZB_fit_d_ebp <- readRDS(save_path)
} else {
  ZB_fit_d_ebp <- glmZiBFit(sy, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  saveRDS(ZB_fit_d_ebp, save_path)
}

th_debp = top_hits(ZB_fit_d_ebp)

```

Down to 97 hits...

```{r}
plot_ani_dist(sy, 'RvsP', th_debp$Contig_name[1:20]) 
```

Not enough to kill *B. animalis* though...

###  Post-hoc testing
```{r}
th_debp_ph = strainspy::comp_ani_diff_and_posthoc_test(sy, ZB_fit, th_debp)

table(th_debp_ph$Comment, useNA = 'always') # now there are only 15 hits, one less from before

plot_ani_dist(sy, 'RvsP', th_debp_ph$Contig_name[is.na(th_debp_ph$Comment)]) 
```

### Common/different hits
```{r}

disp_hits = th_debp_ph$Contig_name[is.na(th_debp_ph$Comment)]
ndisp_hits = th_ph$Contig_name[is.na(th_ph$Comment)]

# Common ones
plot_ani_dist(sy, 'RvsP', intersect(disp_hits,ndisp_hits))

# Only in model without dispersion prior:
plot_ani_dist(sy, 'RvsP', setdiff(ndisp_hits, disp_hits))

# Only in model WITH dispersion prior:
plot_ani_dist(sy, 'RvsP', setdiff(disp_hits, ndisp_hits))


```


