---
title: "Analysis of rUTI metagenomes"
output:
  github_document
date: "2025-09-25"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Run ZiB analysis in q99 mode

## Load dependencies

```{r, warning=FALSE, message=FALSE}
library(SummarizedExperiment)
library(tidyverse)
library(strainspy)
library(dplyr)
library(ggplot2)
library(ggvenn)
library(stringr)
library(readxl)
library(stringdist)
```

## Load metadata 

```{r}
meta <- readRDS("./data/rUTI/meta.rds") 
# Remoe first column
meta = meta[,-1]
# Merged_samples
meta$Run = unname(sapply(meta$Run, function(x) unlist(strsplit(x, ','))[1]))
# set up contrasts/reference levels
meta$Cohort = factor(meta$Cohort, levels = c("Control", "rUTI"))
# set up subjects
meta$UMB = factor(meta$UMB)
# Set up sample type
meta$sample= factor(meta$sample)

```

## Load sylph outputs

### Containment ANI
```{r}
sy <- read_sylph("./data/rUTI/combined_q_99.tsv.gz") # q99)
sy <- filter_by_presence(sy, min_nonzero = 39) # filter at 10%

# annoying renames to match meta V sylph file
colnames(sy) <- gsub("_1", "", colnames(sy))
colData(sy)$Sample_file <- gsub("_1", "", basename(colData(sy)$Sample_file))

dim(sy)

# Checks before merging metadata
all(colnames(sy) %in% meta$Run)
all(meta$run_accession %in% colnames(sy))

sy = modify_metadata(sy, meta)
```

### Abundance
```{r}
sy_ab <- read_sylph("./data/rUTI/combined_p_95.tsv.gz", variable = 'Taxonomic_abundance')
sy_ab <- filter_by_presence(sy_ab, min_nonzero = 39) # filter at 10%

# annoying renames to match meta V sylph file
colnames(sy_ab) <- gsub("_1", "", colnames(sy_ab))
colData(sy_ab)$Sample_file <- gsub("_1", "", basename(colData(sy_ab)$Sample_file))

dim(sy_ab)

# There are two samples missing here
meta_ab = meta[meta$Run %in% sy_ab$Sample_file, ]

# Checks before merging metadata
all(colnames(sy_ab) %in% meta_ab$Run)
all(meta_ab$run_accession %in% colnames(sy_ab))

sy_ab = modify_metadata(sy_ab, meta_ab)
```

# Differences between gut microbiomes

Take out the gut microbiome samples

```{r}
meta_gut = meta[-which(meta$sample == "urine"), ]
sy_gut = sy[,-which(meta$sample == "urine")]

meta_ab_gut = meta_ab[-which(meta_ab$sample == "urine"), ]
sy_ab_gut = sy_ab[,-which(meta_ab$sample == "urine")]
```

## Baseline differences

### Retain only baseline samples

Exclude samples around 2 weeks of a UTI event and/or two weeks after Antibiotic exposure

cANI: 
```{r}
meta_gut_bl = meta_gut

rUTI_idx = which(!meta_gut_bl$Cohort == "Control")

non_baseline_idx = sort(rUTI_idx[unique(c(which(meta_gut_bl$time_since_uti[rUTI_idx] <= 14), 
                                          which(meta_gut_bl$time_before_uti[rUTI_idx] >= -14), 
                                          which(meta_gut_bl$time_since_ab[rUTI_idx] <= 14)))])

meta_gut_bl = meta_gut_bl[-non_baseline_idx, ]

sy_gut_bl = sy_gut[,-non_baseline_idx]
sy_gut_bl = modify_metadata(sy_gut_bl, meta_gut_bl, replace = T)

```

Abundance:
```{r}
meta_ab_gut_bl = meta_ab_gut

rUTI_idx = which(!meta_ab_gut_bl$Cohort == "Control")

non_baseline_idx = sort(rUTI_idx[unique(c(which(meta_ab_gut_bl$time_since_uti[rUTI_idx] <= 14), 
                                          which(meta_ab_gut_bl$time_before_uti[rUTI_idx] >= -14), 
                                          which(meta_ab_gut_bl$time_since_ab[rUTI_idx] <= 14)))])

meta_ab_gut_bl = meta_ab_gut_bl[-non_baseline_idx, ]


sy_ab_gut_bl = sy_ab_gut[,-non_baseline_idx]
sy_ab_gut_bl = modify_metadata(sy_ab_gut_bl, meta_gut_bl, replace = T)

```


### Baseline difference in composition
```{r}
design <- as.formula("~ Cohort + (1 | UMB)")
save_path <- "output_rds/rUTI_zib_q_99_ebp_baseline.rds"
if(file.exists(save_path)){
  ZB_fit_bl <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy_gut_bl, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_bl <- glmZiBFit(sy_gut_bl,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_bl, save_path)
}

top_hits(ZB_fit_bl, method = "BH")
```

No statistically significance difference in strain composition between groups

### Baseline difference in abundance
```{r}
save_path <- "output_rds/rUTI_zib_q_99_ebp_baseline_abundance.rds"
if(file.exists(save_path)){
  abud_fit_bl <- readRDS(save_path)
} else {
  abud_fit_bl <- abundanceFit(sy_ab_gut_bl, design, transform = 'CLR', nthreads = parallel::detectCores())
  saveRDS(abud_fit_bl, save_path)
}

top_hits(abud_fit_bl, method = "BH")
```

No statistically significance difference in strain abundance between groups.

### Load GTDB taxonomy

```{r}
taxonomy <- read_taxonomy("data/TAXONOMY/sylph_DB_taxonomy_99.tsv")
```

### Top hits
```{r}
# Top beta
strainspy:::add_tax2tophits(top_hits(ZB_fit_bl, alpha = 1), taxonomy, c("Family", "Species")) %>%
  group_by(Family) %>%
  slice_min(p_value, n = 1, with_ties = FALSE) %>%  # top species per family
  ungroup() %>%
  arrange(p_value) %>%                               # sort families by their top species' p-value
  slice_head(n = 10) %>%
  select(Family, Species)

# Top Zi
strainspy:::add_tax2tophits(top_hits(ZB_fit_bl, alpha = 1), taxonomy, c("Family", "Species")) %>%
  group_by(Family) %>%
  slice_min(zi_p_value, n = 1, with_ties = FALSE) %>%  # top species per family
  ungroup() %>%
  arrange(p_value) %>%                               # sort families by their top species' p-value
  slice_head(n = 10) %>%
  select(Family, Species)

# Top abundance
strainspy:::add_tax2tophits(top_hits(abud_fit_bl, alpha = 1), taxonomy, c("Family", "Species")) %>%
  group_by(Family) %>%
  slice_min(p_value, n = 1, with_ties = FALSE) %>%  # top species per family
  ungroup() %>%
  arrange(p_value) %>%                               # sort families by their top species' p-value
  slice_head(n = 10) %>%
  select(Family, Species)
```

These don't align with the paper findings much, but the paper results are not corrected for multiple testing...

## Differences during UTI and Antibiotic events

```{r}
meta_gut$UTI = as.factor(meta_gut$UTI_day | meta_gut$UTI_proximal_matched)
meta_gut$AB = meta_gut$time_since_ab <= 21 # Antibiotic exposure within 3 weeks
meta_gut$AB[is.na(meta_gut$AB)] = FALSE
meta_gut$AB = as.factor(meta_gut$AB)

sy_gut = modify_metadata(sy_gut, meta_gut, replace = T)
```

## Fit the full model on all gut microbiomes

```{r}
design <- as.formula("~ Cohort + UTI + AB + (1 | UMB)")

save_path <- "output_rds/rUTI_zib_q_99_ebp.rds"
# Run with ebp - this looks a bit less noisy
if(file.exists(save_path)){
  ZB_fit <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy_gut, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit <- glmZiBFit(sy_gut,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit, save_path)
}

top_hits(ZB_fit, coef = 2, method = "BH")
top_hits(ZB_fit, coef = 4, method = "BH")
```

No real signal detected. For the cohorts (`coef = 2`) is not surprising, given what we saw before, there aren't many significant differences between cohorts.
Antibiotic usage and UTI are strongly correlated, I expected some hits, but nothing comes up.

### Active infection vs. all other samples

```{r, fig.width=12}
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = T, coef = 3, method = 'BH')
plot_manhattan(ZB_fit, taxonomy = taxonomy, aggregate_by_taxa = F, coef = 3) 
plot_volcano(ZB_fit, coef = 3, label = T, method = "BH")

```

Looks like there are hits. Dominated by presence/absence of many Collinsella sp. and a strain replacement of *Ruminococcus bromi*. 


## Differences in gut microbiome around infection compared to baseline in rUTI 
```{r, fig.width=12}

# retain only patients who had an episode of UTI during trial
idx <- unlist(lapply(split(seq_len(nrow(meta_gut)), meta_gut$UMB),
                     function(x) if(!all(is.na(meta_gut$time_since_uti[x]))) x else NULL))

meta_ruti = meta_gut[idx, ]
# containment ANI
sy_gut_ruti = sy_gut[,idx]


sy_gut_ruti = filter_by_presence(sy_gut_ruti, min_nonzero = 20) # re apply 0 filter 

meta_ruti$UTI = "baseline"
meta_ruti$UTI[which(meta_ruti$time_before_uti > -31)] = "pre"
meta_ruti$UTI[which(meta_ruti$time_since_uti < 15)] = "post"

table(meta_ruti$UTI, meta_ruti$AB)

meta_ruti$UTI = factor(meta_ruti$UTI, levels = c("baseline", "pre", "post")) 

sy_gut_ruti = modify_metadata(sy_gut_ruti, meta_ruti, replace = T)
design <- as.formula("~ UTI + AB + (1 | UMB)")

save_path <- "output_rds/rUTI_infection_zib_q_99_ebp.rds"
# Run with ebp - this looks a bit less noisy
if(file.exists(save_path)){
  ZB_fit_inf <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy_gut_ruti, strainspy:::nobars_(design), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_inf2 <- glmZiBFit(sy_gut_ruti,  design, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_inf, save_path)
}

strainspy:::add_tax2tophits(top_hits(ZB_fit_inf, 2, method = "BH" ), taxonomy)

th = strainspy:::add_tax2tophits(top_hits(ZB_fit_inf, 3, method = "BH" ), taxonomy, columns = c("Species", "Genus"))
sort(table(th$Genus), decreasing = T) 

plot_volcano(ZB_fit_inf, coef = 3, method = "BH")


```

There are clear effects of antibiotic exposure post UTI - disappearance and replacement of butyrate produces such as *Blautia wexlerae* and the disappearance of Collinsella sp. and many Escherechia sp.

## E. coli bloom capture pre-UTI
```{r}
sy_gut_ruti_ec = sy_gut_ruti[grep("Escherichia", rownames(sy_gut_ruti)),] # only 899 e coli samples here

meta_ec <- meta_ruti
sy_gut_ruti_ec = sy_gut_ruti_ec[, which(meta_ec$UMB == 8 | meta_ec$UMB == 12 | meta_ec$UMB == 20 | meta_ec$UMB == 23)]
meta_ec = meta_ec[which(meta_ec$UMB == 8 | meta_ec$UMB == 12 | meta_ec$UMB == 20 | meta_ec$UMB == 23), ]

meta_ec$window <- "baseline"

# 2. pre_uti: within X days before UTI OR after UTI but before antibiotics
meta_ec$window[(!is.na(meta_ec$time_before_uti) & abs(meta_ec$time_before_uti) <= 7)] <- "pre_uti"
bl = which(meta_ec$window == "baseline")
meta_ec$window[bl[meta_ec$time_since_ab[bl] < 21 & !is.na(meta_ec$time_since_ab[bl])]] <- "post_ab"

meta_ec$sum_ec = colSums(assay(sy_gut_ruti_ec))

meta_ec %>% group_by(UMB) %>%
  summarise(sum(window == "pre_uti"))

ggplot(meta_ec, aes(x = window, y = sum_ec, fill = window)) + 
    geom_boxplot(outlier.shape = 21, alpha = 0.7) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 1, color = "black") +
    theme_minimal(base_size = 14) +
    labs(
        x = "Window",
        y = "Total ANI signal (sum across 899 strains)",
        title = "Total E. coli ANI per sample across windows"
    ) +
    theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5)
    ) + 
  facet_wrap(~UMB)

design_ec = as.formula('~window + (1 | UMB)')
sy_gut_ruti_ec = modify_metadata(sy_gut_ruti_ec, meta_ec, replace = T)

save_path <- "output_rds/rUTI_infection_zib_q_99_ebp_ecoli.rds"
# Run with ebp - this looks a bit less noisy
if(file.exists(save_path)){
  ZB_fit_ecoli <- readRDS(save_path)
} else {
  ebp = compute_eb_priors(sy_gut_ruti_ec, strainspy:::nobars_(design_ec), nthreads = parallel::detectCores(),low_cutoff = 0, high_cutoff = Inf)
  ZB_fit_ecoli <- glmZiBFit(sy_gut_ruti_ec, design_ec, MAP_prior = ebp, nthreads = parallel::detectCores())
  saveRDS(ZB_fit_ecoli, save_path)
}

```

## Closer investigation of the strains that disappear post infection
```{r}
th_inf = strainspy:::add_tax2tophits(top_hits(ZB_fit_inf, coef = 3, method = "BH"), taxonomy, c("Family", "Species"))
# Mostly driven by E coli!
head(sort(table(th_inf$Species), decreasing = T))

th_inf_ec = th_inf[th_inf$Species == "Escherichia coli", ]

# https://doi.org/10.1101/2025.04.13.648558 has a nice supplementary table classifying E coli strains - could be helpful
m1 = readxl::read_xlsx("./data/rUTI/media-1.xlsx", sheet = 2)
# Strain names don't match easily, let's try a few things - chatGPT solution
# --- 1. Normalize accessions: GCF -> GCA ---
th_inf_ec$Genome_file_norm <- sub("^GCF_", "GCA_", th_inf_ec$Genome_file)
m1$Accession_norm      <- sub("^GCF_", "GCA_", m1$`Accession ID`)

# --- 2. Direct accession match ---
th_inf_ec$direct_match_idx <- match(th_inf_ec$Genome_file_norm, m1$Accession_norm)

# --- 3. Clean strain names for fuzzy matching ---
clean_name <- function(x){
  x <- tolower(x)
  x <- gsub("chromosome|whole genome shotgun sequence|contig[0-9]+|_000[0-9]+|NODE_[0-9]+.*", "", x)
  x <- gsub("[^a-z0-9]", "", x)
  trimws(x)
}

th_inf_ec$strain_clean <- clean_name(th_inf_ec$Contig_name)
m1$strain_clean     <- clean_name(m1$Strain)

# --- 4. Fuzzy match for rows without direct accession match ---
no_match_idx <- which(is.na(th_inf_ec$direct_match_idx))
th_inf_ec$fuzzy_match_idx <- NA

if(length(no_match_idx) > 0){
  fuzzy_idx <- amatch(th_inf_ec$strain_clean[no_match_idx],
                      m1$strain_clean,
                      maxDist = 5)   # adjust tolerance
  th_inf_ec$fuzzy_match_idx[no_match_idx] <- fuzzy_idx
}

# --- 5. Combine indices ---
th_inf_ec$final_idx <- ifelse(!is.na(th_inf_ec$direct_match_idx),
                              th_inf_ec$direct_match_idx,
                              th_inf_ec$fuzzy_match_idx)

# --- 6. Pull only Strain and Pathogenicity type ---
th_inf_ec$Strain <- m1$Strain[th_inf_ec$final_idx]
th_inf_ec$Pathogenicity_type <- m1$`Pathogenicity type`[th_inf_ec$final_idx]
sort(table(th_inf_ec$Pathogenicity_type, useNA = 'always'), decreasing = T)

```

Antibiotic exposure depletes a substantial proportion of the *E. coli* population, including a large number of commensals. 



## Species that change in Abundance
```{r}
sy_ab_gut_ruti = sy_ab_gut[,idx]
sy_ab_gut_ruti = modify_metadata(sy_ab_gut_ruti, meta_ruti, replace = T)

save_path <- "output_rds/rUTI_infection_zib_q_99_abud.rds"
if(file.exists(save_path)){
  abud_fit_inf <- readRDS(save_path)
} else {
  abud_fit_inf <- abundanceFit(sy_ab_gut_ruti, design, transform = 'CLR', nthreads = parallel::detectCores())
  saveRDS(abud_fit_inf, save_path)
}

strainspy:::add_tax2tophits(top_hits(abud_fit_inf, coef = 2, method = "BH"), taxonomy)
strainspy:::add_tax2tophits(top_hits(abud_fit_inf, coef = 3, method = "BH"), taxonomy)
strainspy:::add_tax2tophits(top_hits(abud_fit_inf, coef = 4, method = "BH"), taxonomy)

```

