---
title: "Simulations with spiked P. dis strain analysed using Sylph Query"
output: github_document
date: "2026-04-15"
editor_options:
  chunk_output_type: console
---

### Load data and prepare for testing

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(strainspy)
library(ggplot2)
library(dplyr)
library(ggrepel)
# Queries
d_q99 <- read_sylph("p_distasonis_sims/pdis_q99.tsv")

# spiked samples
ss = readLines("p_distasonis_sims/spike_genome_ids.txt") # These are the baseline samples, not spiked
ss_spikes = c(paste(ss, "_sx1", sep = ""), paste(ss, "_sx3", sep = ""), paste(ss, "_sx5", sep = "")) # These are the spikes

# Remove _1 annoyance
colnames(d_q99) = sub("_1", "", colnames(d_q99))
d_q99@colData$Sample_file = sub("_1", "", colnames(d_q99))

# add the metadata
d_q99 = strainspy:::modify_metadata(se = d_q99, meta_data = data.frame(run_accession = colnames(d_q99),
                                                                       spiked = as.factor(sapply(colnames(d_q99), function(x) x%in% c(ss, ss_spikes)))))

# I am marking the baseline unspiked ones as spiked to show the null doesn't detect anything

d_q99 <- filter_by_presence(d_q99, min_nonzero = 25)

# taxonomies
tax_99 = read_taxonomy("data/TAXONOMY/sylph_DB_taxonomy_99.tsv")

design <- as.formula(" ~ spiked")

```

## Fit ZIB model

### query 99, baseline (no signal)
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# This simulation does not have any signal, we shouldn't detect anything!

# Subset
dim(d_q99)
rm_idx = grep("_sx", colnames(d_q99))
d_q99_test = d_q99[,-rm_idx]
dim(d_q99_test)

# These TRUE = 20 are NOT spiked in, this is just to show there is no detectable signal
table(d_q99_test@colData$spiked) 


save_path = "output_rds/spike_pdis_zib_q_99_bl.rds"
if(file.exists(save_path)){
  fit_zib_99_bl <- readRDS(save_path)
} else {
  system.time({
    ebp = compute_eb_priors(d_q99_test, design, nthreads = parallel::detectCores())
    fit_zib_99_bl <- glmZiBFit(d_q99_test, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  })
  saveRDS(fit_zib_99_bl, save_path)
}

top_hits(fit_zib_99_bl) # no hits as expected
```

### query 99, cov1x

```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# baseline false samples and higher coverage ones
rm_idx = c(sapply(ss, function(x) which(colnames(d_q99) %in% x)), grep("_sx3", colnames(d_q99)), grep("_sx5", colnames(d_q99)))
d_q99_test = d_q99[,-rm_idx]
dim(d_q99_test)

# These 20 are spiked ata coverage of 1x
table(d_q99_test@colData$spiked) 

save_path = "output_rds/spike_pdis_zib_q_99_cov1x.rds"
if(file.exists(save_path)){
  fit_zib_99_c1 <- readRDS(save_path)
} else {
  system.time({
    ebp = compute_eb_priors(d_q99_test, design, nthreads = parallel::detectCores())
    fit_zib_99_c1 <- glmZiBFit(d_q99_test, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  })
  saveRDS(fit_zib_99_c1, save_path)
}

th_c1 = strainspy:::add_tax2tophits(top_hits(fit_zib_99_c1), tax_99)

# All 76 hits from the same species
table(th_c1$Species)

plot_manhattan(fit_zib_99_c1, taxonomy = tax_99, method = "BH", tax_levels = c("Order", "Phylum", "Genus", "Species"))
plot_manhattan(fit_zib_99_c1, taxonomy = tax_99, aggregate_by_taxa = F)
plot_volcano(fit_zib_99_c1, label = T)

plot_ani_dist(d_q99, phenotype = 'spiked', contigs = th_c1$Contig_name[1:5], contig_names = th_c1$Genome_file[1:5])
# Parabacteroides distasonis is in all samples, so, this can only be a beta hit!
# GCF_024791025.1 is the top hit - it is the spiked strain. And for the spiked samples, the ANI is 100%


```

### query 99, cov3x

```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# baseline false samples and higher coverage ones
rm_idx = c(sapply(ss, function(x) which(colnames(d_q99) %in% x)), grep("_sx1", colnames(d_q99)), grep("_sx5", colnames(d_q99)))
d_q99_test = d_q99[,-rm_idx]
dim(d_q99_test)

# These 20 are spiked ata coverage of 3x
table(d_q99_test@colData$spiked) 

save_path = "output_rds/spike_pdis_zib_q_99_cov3x.rds"
if(file.exists(save_path)){
  fit_zib_99_c3 <- readRDS(save_path)
} else {
  system.time({
    ebp = compute_eb_priors(d_q99_test, design, nthreads = parallel::detectCores())
    fit_zib_99_c3 <- glmZiBFit(d_q99_test, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  })
  saveRDS(fit_zib_99_c3, save_path)
}

th_c3 = strainspy:::add_tax2tophits(top_hits(fit_zib_99_c3), tax_99)

# All 91 hits from the same species
table(th_c3$Species)

plot_manhattan(fit_zib_99_c3, taxonomy = tax_99, method = "BH", tax_levels = c("Order", "Phylum", "Genus", "Species"))
plot_manhattan(fit_zib_99_c3, taxonomy = tax_99, aggregate_by_taxa = F)
plot_volcano(fit_zib_99_c3, label = T)

plot_ani_dist(d_q99, phenotype = 'spiked', contigs = th_c3$Contig_name[1:5], contig_names = th_c3$Genome_file[1:5])
# Parabacteroides distasonis is in all samples, so, this can only be a beta hit!
# GCF_024791025.1 is the top hit - it is the spiked strain. And for the spiked samples, the ANI is 100%


```

### query 99, cov5x

```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# baseline false samples and higher coverage ones
rm_idx = c(sapply(ss, function(x) which(colnames(d_q99) %in% x)), grep("_sx1", colnames(d_q99)), grep("_sx3", colnames(d_q99)))
d_q99_test = d_q99[,-rm_idx]
dim(d_q99_test)

# These 20 are spiked ata coverage of 3x
table(d_q99_test@colData$spiked) 

save_path = "output_rds/spike_pdis_zib_q_99_cov5x.rds"
if(file.exists(save_path)){
  fit_zib_99_c5 <- readRDS(save_path)
} else {
  system.time({
    ebp = compute_eb_priors(d_q99_test, design, nthreads = parallel::detectCores())
    fit_zib_99_c5 <- glmZiBFit(d_q99_test, design, nthreads = parallel::detectCores(), MAP_prior = ebp)
  })
  saveRDS(fit_zib_99_c5, save_path)
}

th_c5 = strainspy:::add_tax2tophits(top_hits(fit_zib_99_c5), tax_99)

# All 91 hits from the same species
table(th_c5$Species)

plot_manhattan(fit_zib_99_c5, taxonomy = tax_99, method = "BH", tax_levels = c("Order", "Phylum", "Genus", "Species"))
plot_manhattan(fit_zib_99_c5, taxonomy = tax_99, aggregate_by_taxa = F)
plot_volcano(fit_zib_99_c5, label = T)

plot_ani_dist(d_q99, phenotype = 'spiked', contigs = th_c5$Contig_name[1:5], contig_names = th_c5$Genome_file[1:5])
# Parabacteroides distasonis is in all samples, so, this can only be a beta hit!
# GCF_024791025.1 is the top hit - it is the spiked strain. And for the spiked samples, the ANI is 100%


```

StrainSpy works regardless of coverage, now to test AnPan


### Load data and prepare for testing
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}
library(anpan)
library(ape)
library(tibble)
tree = read.tree("p_distasonis_sims/spiked_tree.tre")

metadata = tibble(sample_id = tree$tip.label,
                  spiked   = as.factor(sapply(tree$tip.label, function(x) x%in% c(ss, sub("_", "_1_", paste(ss_spikes, ".fastq", sep = "")) ))) )

table(metadata$spiked)

plot_outcome_tree(tree,
                  metadata, 
                  covariates = c(),
                  outcome    = "spiked")

```

Looks like the left-most branch has a bunch of spiked ones, it'll be interesting to see variation with coverage

## anPan PGLMM

### baseline (no signal)
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# This simulation does not have any signal, we shouldn't detect anything!

# Subset
rm_idx = grep("_sx", metadata$sample_id)

metadata_test = metadata[-rm_idx, ]
table(metadata_test$spiked)
# Only 76/100 samples had P. dist at a detectable level for strainphlan
# These TRUE = 12 are NOT spiked in, this is just to show there is no detectable signal

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
plot_outcome_tree(tree_test,
                  metadata_test, 
                  covariates = c(),
                  outcome    = "spiked")

# Clearly no signal

save_path = "output_rds/spike_pdis_anpan_bl.rds"
if(file.exists(save_path)){
  anpan_res_bl <- readRDS(save_path)
} else {
  system.time({
    anpan_res_bl = anpan_pglmm(meta_file       = metadata_test,
                               tree_file       = tree_test,
                               outcome         = "spiked",
                               family          = "binomial",
                               bug_name        = "Pdis",
                               reg_noise       = TRUE,
                               loo_comparison  = TRUE,
                               run_diagnostics = FALSE,
                               refresh         = 500,
                               show_plot_tree  = FALSE,
                               show_post       = FALSE)
  })
  saveRDS(anpan_res_bl, save_path)
}

anpan_res_bl$loo$comparison

```

Clearly no signal - confirmed by PGLMM

### cov1x
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}
# Subset
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx3", metadata$sample_id), grep("_sx5", metadata$sample_id)))

metadata_test = metadata[-rm_idx, ]
table(metadata_test$spiked)
# Only 20/20 spiked samples are now detected, but only non spiked are present - should be plenty 

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
plot_outcome_tree(tree_test,
                  metadata_test, 
                  covariates = c(),
                  outcome    = "spiked")

# Signal is a bit weak

save_path = "output_rds/spike_pdis_anpan_cov1x.rds"
if(file.exists(save_path)){
  anpan_res_c1 <- readRDS(save_path)
} else {
  system.time({
    anpan_res_c1 = anpan_pglmm(meta_file       = metadata_test,
                               tree_file       = tree_test,
                               outcome         = "spiked",
                               family          = "binomial",
                               bug_name        = "Pdis",
                               reg_noise       = TRUE,
                               loo_comparison  = TRUE,
                               run_diagnostics = FALSE,
                               refresh         = 500,
                               show_plot_tree  = FALSE,
                               show_post       = FALSE)
  })
  saveRDS(anpan_res_c1, save_path)
}

anpan_res_c1$loo$comparison

```

The phylogenetic model seems to fit better, but the difference doesn't seem clear (less than 2 standard errors difference in ELPD).

### cov3x
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}

# Subset
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx1", metadata$sample_id), grep("_sx5", metadata$sample_id)))

metadata_test = metadata[-rm_idx, ]
table(metadata_test$spiked)
# Only 20/20 spiked samples are now detected, but only non spiked are present - should be plenty 

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
plot_outcome_tree(tree_test,
                  metadata_test, 
                  covariates = c(),
                  outcome    = "spiked")

# Clearly signal is present

save_path = "output_rds/spike_pdis_anpan_cov3x.rds"
if(file.exists(save_path)){
  anpan_res_c3 <- readRDS(save_path)
} else {
  system.time({
    anpan_res_c3 = anpan_pglmm(meta_file       = metadata_test,
                               tree_file       = tree_test,
                               outcome         = "spiked",
                               family          = "binomial",
                               bug_name        = "Pdis",
                               reg_noise       = TRUE,
                               loo_comparison  = TRUE,
                               run_diagnostics = FALSE,
                               refresh         = 500,
                               show_plot_tree  = FALSE,
                               show_post       = FALSE)
  })
  saveRDS(anpan_res_c3, save_path)
}

anpan_res_c3$loo$comparison

```

The phylogenetic model seems to fit better, and the difference seems clear (more than 2 standard errors difference in ELPD).

### cov5x
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}
# Subset
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx1", metadata$sample_id), grep("_sx3", metadata$sample_id)))

metadata_test = metadata[-rm_idx, ]
table(metadata_test$spiked)
# Only 20/20 spiked samples are now detected, but only non spiked are present - should be plenty 

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
plot_outcome_tree(tree_test,
                  metadata_test, 
                  covariates = c(),
                  outcome    = "spiked")

# Clearly dominant signal

save_path = "output_rds/spike_pdis_anpan_cov5x.rds"
if(file.exists(save_path)){
  anpan_res_c5 <- readRDS(save_path)
} else {
  system.time({
    anpan_res_c5 = anpan_pglmm(meta_file       = metadata_test,
                               tree_file       = tree_test,
                               outcome         = "spiked",
                               family          = "binomial",
                               bug_name        = "Pdis",
                               reg_noise       = TRUE,
                               loo_comparison  = TRUE,
                               run_diagnostics = FALSE,
                               refresh         = 500,
                               show_plot_tree  = FALSE,
                               show_post       = FALSE)
  })
  saveRDS(anpan_res_c5, save_path)
}

anpan_res_c5$loo$comparison

```

The phylogenetic model seems to fit better, and the difference seems clear (more than 2 standard errors difference in ELPD).


AnPan works as well - when there is signal in the phylogeny, it can detect it accurately. However, building the tree can be tricky if the strain coverage is low. StrainSpy can work equally well in the lower coverage simulations as well, without requiring a phylogeny. 

## Figure for paper
```{r, fig.width=12, fig.asp=0.75, out.width='100%'}
# strainspy result

spike = "GCF_024791025.1"

th_plt = rbind(cbind(th_c1, coverage = "x1"),
               cbind(th_c3, coverage = "x3"),
               cbind(th_c5, coverage =  "x5"))

th_plt[nrow(th_plt)+1, ] = th_plt[nrow(th_plt),]
# manually add the baseline hits (there are none!)
th_plt$coefficient[nrow(th_plt)] = 0 
th_plt$p_adjust[nrow(th_plt)] = 1
th_plt$coverage[nrow(th_plt)] = "Null"
th_plt$Contig_name[nrow(th_plt)] = ""


th_plt$neglog_p = -log10(th_plt$p_adjust)

label_df <- th_plt %>%
  group_by(coverage) %>%
  arrange(p_adjust, .by_group = TRUE) %>%
  slice_head(n = 5) %>%
  ungroup()

# No label needed for the fake point
rmidx = which(label_df$coverage == "Null")
if(length(rmidx) > 0) label_df = label_df[-rmidx, ]

top1 <- th_plt %>%
  group_by(coverage) %>%
  arrange(p_adjust) %>%
  slice(1)

ggplot(th_plt, aes(x = coefficient, y = neglog_p)) +
  geom_point(
    data = subset(th_plt, Genome_file != spike),
    aes(color = "other"),
    size = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(th_plt, Genome_file == spike),
    color = "red",
    size = 2.5,
    stroke = 1.2
  ) +
  geom_text_repel(
    data = label_df,
    aes(label = Genome_file),
    size = 4.5,
    fontface = "bold",
    box.padding = 0.4,
    point.padding = 0.3,
    segment.color = "grey50",
    max.overlaps = Inf
  ) + 
  
  scale_color_manual(values = c("grey70", "red")) +
  facet_wrap(~coverage, nrow = 1) +
  theme_classic() +
  labs(
    x = "Effect size (coefficient)",
    y = expression(-log[10](adjusted~p))
  ) +
  guides(color = "none") + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.3) +
  theme_classic(base_size = 16) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1, "lines")
  )


# anpan result

## Null model
rm_idx = rm_idx = grep("_sx", metadata$sample_id)
metadata_test = metadata[-rm_idx, ]

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
tp_null = plot_outcome_tree(tree_test,
                        metadata_test, 
                        covariates = c(),
                        outcome    = "spiked", return_tree_df = T)

tp_null$tree_plot + theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

## Coverage 1 
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx3", metadata$sample_id), grep("_sx5", metadata$sample_id)))
metadata_test = metadata[-rm_idx, ]

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
tp1 = plot_outcome_tree(tree_test,
                        metadata_test, 
                        covariates = c(),
                        outcome    = "spiked", return_tree_df = T)

tp1$tree_plot + theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

## Coverage 3 
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx1", metadata$sample_id), grep("_sx5", metadata$sample_id)))
metadata_test = metadata[-rm_idx, ]

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
tp3 = plot_outcome_tree(tree_test,
                        metadata_test, 
                        covariates = c(),
                        outcome    = "spiked", return_tree_df = T)

tp3$tree_plot + theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

## Coverage 5
rm_idx = unlist(c(sapply(ss, function(x) which(metadata$sample_id %in% x)), grep("_sx1", metadata$sample_id), grep("_sx3", metadata$sample_id)))
metadata_test = metadata[-rm_idx, ]

# Drop the same leaves from the tree
tree_test = drop.tip(tree, metadata$sample_id[rm_idx])
tp5 = plot_outcome_tree(tree_test,
                        metadata_test, 
                        covariates = c(),
                        outcome    = "spiked", return_tree_df = T)

tp5$tree_plot + theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Results summary plot
anpan_summary <- tibble(
  coverage = c("Null", "x1", "x3", "x5"),
  
  elpd_diff = -c(
    anpan_res_bl$loo$comparison["pglmm_fit", "elpd_diff"],
    anpan_res_c1$loo$comparison["base_fit", "elpd_diff"],
    anpan_res_c3$loo$comparison["base_fit", "elpd_diff"],
    anpan_res_c5$loo$comparison["base_fit", "elpd_diff"]
  ),
  
  se_diff = c(
    anpan_res_bl$loo$comparison["pglmm_fit", "se_diff"],
    anpan_res_c1$loo$comparison["base_fit", "se_diff"],
    anpan_res_c3$loo$comparison["base_fit", "se_diff"],
    anpan_res_c5$loo$comparison["base_fit", "se_diff"]
  )
) %>%
  mutate(
    abs_elpd = abs(elpd_diff),
    z_score = abs_elpd / se_diff
  )

anpan_summary$coverage <- factor(anpan_summary$coverage,
                                 levels = c("Null", "x1","x3","x5"))

anpan_summary <- anpan_summary %>%
  mutate(
    x = as.numeric(coverage),
    signal = abs(elpd_diff) / se_diff
  )

ggplot(anpan_summary, aes(x = x, y = elpd_diff, group = 1)) +
  
  # uncertainty
  geom_errorbar(
    aes(
      ymin = elpd_diff - se_diff,
      ymax = elpd_diff + se_diff
    ),
    width = 0.1
  ) +
  
  # trajectory
  geom_line(linewidth = 1) +
  
  # points
  geom_point(size = 4) +
  
  # reference line
  geom_hline(yintercept = 0, linetype = "dashed") +
  
  # threshold (your rule)
  geom_hline(yintercept = 2, linetype = "dotted", color = "red") +
  
  scale_x_continuous(
    breaks = 1:4,
    labels = c("Null", "1x","3x","5x")
  ) +
  
  theme_classic(base_size = 16) +
  
  labs(
    x = "Coverage",
    y = expression(Delta~ELPD~abs(phylogeny - base))
  )

```