---
title: "Simulations"
output:
  html_document: default
date: "2025-02-04"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Download data

Sample 200 sample IDs from the Zeevi et al., dataset (PRJEB11532)

```
{ head -n 1 PRJEB11532_ena_meta.tsv; tail -n +2 PRJEB11532_ena_meta.tsv | shuf -n 200; } > subsample200_PRJEB11532_ena_meta.tsv
cut -f1 subsample200_PRJEB11532_ena_meta.tsv > subsample200_PRJEB11532_ids.tsv


kingfisher get --run-identifiers-list subsample200_PRJEB11532_ids.tsv --download-threads 5 -m ena-ascp ena-ftp aws-http prefetch
```

### Run Sylph

Run Sylph (v0.8.0) in 'query' mode using the GTDB 99% identity dataset.

```
sylph query gtdb_ordered_99.syldb -u --read-seq-id 99.9 -t 100  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./zeevi_PRJEB11532_query_gtdb_220_id99.tsv ./ERR*[0-9][0-9].fastq.gz
```

```
sylph query /data1/gerryt/sylph_analyses/gtdb_98_ordered_DB/gtdb_ordered_98.syldb -u --read-seq-id 99.9 -t 25  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./zeevi_PRJEB11532_query_gtdb_220_id98.tsv ./ERR*[0-9][0-9].fastq.gz
```

```
sylph profile /data1/gerryt/sylph_analyses/gtdb_98_ordered_DB/gtdb_ordered_98.syldb -u --read-seq-id 99.9 -t 25  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./zeevi_PRJEB11532_profile_gtdb_220_id98.tsv ./ERR*[0-9][0-9].fastq.gz
```

```
sylph profile /data1/gerryt/sylph_analyses/gtdb-r220-c200-dbv1.syldb -u --read-seq-id 99.9 -t 25  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./zeevi_PRJEB11532_profile_gtdb_220_id95.tsv ./ERR*[0-9][0-9].fastq.gz
```

```
sylph query /data1/gerryt/sylph_analyses/gtdb-r220-c200-dbv1.syldb -u --read-seq-id 99.9 -t 25  -1 ./*_1.fastq.gz -2 ./*_2.fastq.gz -o ./zeevi_PRJEB11532_query_gtdb_220_id95.tsv ./ERR*[0-9][0-9].fastq.gz
```


### Function to generate sensitivity heatmaps
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)

gg_facet_plot <- function(op, title = NULL){
  if('sf' %in% colnames(op)){
    op$var = as.numeric(op$sf)
    xlab_ = "Scale Factor"
  } else if('z_prop' %in% colnames(op)){
    op$var = as.numeric(op$z_prop)
    xlab_ = "Zero Proportion"
  }
  
  
  op$Test <- factor(op$Test)
  op$tp = as.numeric(op$tp)
  op$fn = as.numeric(op$fn)
  op$fp = as.numeric(op$fp)
  op$tn = as.numeric(op$tn)
  
  op <- op %>% mutate(
    Sensitivity = tp / (tp + fn) * 100,
    Specificity = tn / (tn + fn) * 100,
    Precision   = tp / (tp + fp) * 100,
    # Recall      = tp / (tp + fn) * 100,
    F1          = 2 * tp / (2 * tp + fp + fn) * 100,
    # MCC         = ifelse(
    #   (tp + fp) * (tp + fn) * (tn + fp) * (tn + fn) == 0,
    #   NA,
    #   ((tp * tn) - (fp * fn)) / sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
    # ) * 100
  )
  
  long_op <- op %>%
    pivot_longer(
      cols = c(Sensitivity, Specificity, Precision, F1),
      names_to = "Metric",
      values_to = "Value"
    )
  
  long_op$Metric <- factor(long_op$Metric, levels = c("Sensitivity", "Specificity", "Precision","F1"))
  
  p = ggplot(long_op, aes(x = var, y = Value, col = Test)) +
    geom_point() +
    geom_line() +
    facet_wrap(~Metric, scales = "free_y") +
    theme_minimal() +
    theme(text = element_text(size = 16)) +
    labs(
      title = title,
      x = xlab_,
      y = NULL
    )
  
  print(p)
  
  
}


gg_heatmaps <- function(op){
  op$sf = as.factor(as.numeric(op$sf))
  op$z_prop = as.factor(as.numeric(op$z_prop))
  op$Test <- factor(op$Test)
  op$tp = as.numeric(op$tp)
  op$fn = as.numeric(op$fn)
  op$fp = as.numeric(op$fp)
  op$tn = as.numeric(op$tn)
  
  op <- op %>% mutate(
    Sensitivity = tp / (tp + fn) * 100,
    Specificity = tn / (tn + fn) * 100,
    Precision   = tp / (tp + fp) * 100,
    # Recall      = tp / (tp + fn) * 100,
    F1          = 2 * tp / (2 * tp + fp + fn) * 100,
    # MCC         = ifelse(
    #   (tp + fp) * (tp + fn) * (tn + fp) * (tn + fn) == 0,
    #   NA,
    #   ((tp * tn) - (fp * fn)) / sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
    # ) * 100
  )
  
  
  op_long <- op %>%
    pivot_longer(cols = c(Sensitivity, Specificity, Precision, F1),
                 names_to = "Metric", values_to = "Value")
  
  op_long$Metric <- factor(op_long$Metric, levels = c("Sensitivity", "Specificity", "Precision", "F1"))
  
  ggplot(op_long, aes(x = z_prop, y = sf, fill = Value)) +
    geom_tile() +
    facet_grid(Metric ~ Test) +
    scale_fill_viridis_c(name = "Value") +
    labs(x = "Zero Proportion", y = "Scale Factor") +
    theme_minimal() +
    theme(
      strip.text = element_text(size = 10, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  
  # ggplot(op, aes(x = z_prop, y = sf, fill = Specificity)) +
  #   geom_tile() +
  #   facet_wrap(~ Test) +  # Separate plots for each test
  #   
  #   scale_fill_viridis_c(name = "Sensitivity") +  # Viridis color scale
  #   
  #   labs(
  #     x = "Zero Proportion",
  #     y = "Scale Factor",
  #     # title = "Sensitivity Across Tests"
  #   ) +
  #   theme_minimal() +
  #   theme(
  #     strip.text = element_text(size = 12, face = "bold"),  # Styling facet labels
  #     axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels if needed
  #   )
  # 
  # ggplot(op, aes(x = zero_prop, y = scale_factor, fill = fp)) +
  #   geom_tile() +
  #   scale_fill_viridis_c(name = "False Positives") +  # Viridis color scale
  #   facet_wrap(~ test) +  # Separate plots for each test
  #   labs(
  #     x = "Zero Proportion",
  #     y = "Scale Factor",
  #     # title = "Sensitivity Across Tests"
  #   ) +
  #   theme_minimal() +
  #   theme(
  #     strip.text = element_text(size = 12, face = "bold"),  # Styling facet labels
  #     axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels if needed
  #   )
  
}

```
### Load data and prepare for simulations

```{r}
library(strainspy)

# Load Zeevi the GTDB 99% identity dataset
# data <- read_sylph("~/Documents/strainspy_manuscript/zeevi/zeevi_PRJEB11532_query_gtdb_220_id99.tsv.gz")
data <- read_sylph("TEST_DATA/zeevi/zeevi_PRJEB11532_query_gtdb_220_id99.tsv.gz")
data <- filter_by_presence(data, min_nonzero = 150)

# Get the ids of cases after a random split
set.seed(789)
cascon = rbinom(round(SummarizedExperiment::ncol(data)),  1, 0.5)
cases = which(cascon == 1)

# setup metadata
meta_data = data.frame(run_accession = colnames(data),
                       Case_status = cascon)

meta_data$Case_status <- factor(meta_data$Case_status)
data = strainspy:::modify_metadata(se = data, meta_data = meta_data)
```

## Sensitivity Analysis
```{r}
# randomly select 100 features (contigs) to rescale abundance - ensure omit NA from cascon model
fit_cascon_null = readRDS("TEST_DATA/zeevi/TEMP_FITS/null_cascon.rds")
set.seed(789)
sel = sample(which(!is.na(fit_cascon_null@coefficients$X.Intercept.)), 100)
contigs = data@elementMetadata$Contig_name[sel]

# fit the 3 models using update option
fit_ob_null = readRDS("TEST_DATA/zeevi/TEMP_FITS/null_ob.rds")
fit_zib_null = readRDS("TEST_DATA/zeevi/TEMP_FITS/null_zib.rds")

```

### Simulate changes in identity
```{r}
save_path = "output_rds/zeevi_sim_id.rds"
if(file.exists(save_path)){
  op_id = readRDS(save_path)
} else {
  # run through by varying beta
  scale_factors = c(seq(0.95, 0.995, by = 0.005),0.999,1)
  
  op_id = data.frame()
  for(scale_factor in scale_factors){
    exp_beta = c()
    exp_zi = c()
    se = data # NULL
    
    data_matrix <- SummarizedExperiment::assay(se) # NULL
    vals_to_mod = data_matrix[sel, cases]
    
    for(i in 1:length(sel)){ 
      tmp = rescale_beta(x = data_matrix[sel[i], cases]/100,
                         beta = scale_factor,
                         zi = 0)
      
      vals_to_mod[i,] = tmp$rescaled*100
      exp_beta[i] = tmp$expected_beta
      exp_zi[i] = tmp$expected_zi
    }
    
    # add modified counts for cases
    data_matrix[sel, cases] = vals_to_mod
    
    # add modified assay back to se
    SummarizedExperiment::assay(se) <- data_matrix
    
    fit_zib_updated = strainspy:::update_fit(fit = fit_zib_null, se = se, update_idx = sel, nthreads = 10)
    fit_ob_updated = strainspy:::update_fit(fit = fit_ob_null, se = se, update_idx = sel, nthreads = 10)
    fit_cascon_updated = strainspy:::update_fit(fit = fit_cascon_null, se = se, update_idx = sel, nthreads = 10, min_identity = 0.9)
    cat(scale_factor, "\n")
    
    op_id = rbind(op_id, 
                  cbind(scale_factor, 
                        rbind(
                          c("zib", strainspy:::get_confusion_mx(top_hits = top_hits(fit_zib_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                          c("ob", strainspy:::get_confusion_mx(top_hits = top_hits(fit_ob_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                          c("cc", strainspy:::get_confusion_mx(top_hits = top_hits(fit_cascon_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)))))
  }
  
  colnames(op_id) = c("sf", "Test", "tp", "fn", "fp", "tn")
  saveRDS(op_id, save_path)
  
}

gg_facet_plot(op_id, "Variation in feature identity")
```

### Simulate changes in presence/absence
```{r}
save_path = "output_rds/zeevi_sim_pa.rds"
if(file.exists(save_path)){
  op_pa = readRDS(save_path)
} else {
  zero_props = seq(0.2, 0.72, by = 0.05)
  op_pa = data.frame()
  for(zero_prop in zero_props){
    exp_beta = c()
    exp_zi = c()
    se = data # NULL
    
    data_matrix <- SummarizedExperiment::assay(se) # NULL
    vals_to_mod = data_matrix[sel, cases]
    
    for(i in 1:length(sel)){ 
      tmp = rescale_beta(x = data_matrix[sel[i], cases]/100,
                         beta = 1,
                         zi = zero_prop)
      
      vals_to_mod[i,] = tmp$rescaled*100
      exp_beta[i] = tmp$expected_beta
      exp_zi[i] = tmp$expected_zi
    }
    
    # add modified counts for cases
    data_matrix[sel, cases] = vals_to_mod
    
    # add modified assay back to se
    SummarizedExperiment::assay(se) <- data_matrix
    
    fit_zib_updated = strainspy:::update_fit(fit = fit_zib_null, se = se, update_idx = sel, nthreads = 10)
    fit_ob_updated = strainspy:::update_fit(fit = fit_ob_null, se = se, update_idx = sel, nthreads = 10)
    fit_cascon_updated = strainspy:::update_fit(fit = fit_cascon_null, se = se, update_idx = sel, nthreads = 10, min_identity = 0.9)
    cat(zero_prop, "\n")
    
    op_pa = rbind(op_pa, 
                  cbind(zero_prop, 
                        rbind(
                          c("zib", strainspy:::get_confusion_mx(top_hits = top_hits(fit_zib_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                          c("ob", strainspy:::get_confusion_mx(top_hits = top_hits(fit_ob_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                          c("cc", strainspy:::get_confusion_mx(top_hits = top_hits(fit_cascon_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)))))
  }
  
  colnames(op_pa) = c("z_prop", "Test", "tp", "fn", "fp", "tn")
  saveRDS(op_pa, save_path)
}

gg_facet_plot(op_pa, "Variation in feature presence/absence")

```


### Simulate both identity and presence/absence changes
```{r}
save_path = "output_rds/zeevi_sim_full.rds"
if(file.exists(save_path)){
  op_full = readRDS(save_path)
} else {
  zero_props = seq(0, 0.72, by = 0.05)
  scale_factors = c(seq(0.95, 0.995, by = 0.005),0.999,1)
  op_full = data.frame()
  
  for(scale_factor in scale_factors){
    for(zero_prop in zero_props){
      
      t0 = Sys.time()
      
      exp_beta = c()
      exp_zi = c()
      se = data # NULL
      
      data_matrix <- SummarizedExperiment::assay(se) # NULL
      vals_to_mod = data_matrix[sel, cases]
      
      for(i in 1:length(sel)){ 
        tmp = rescale_beta(x = data_matrix[sel[i], cases]/100,
                           beta = scale_factor,
                           zi = zero_prop)
        
        vals_to_mod[i,] = tmp$rescaled*100
        exp_beta[i] = tmp$expected_beta
        exp_zi[i] = tmp$expected_zi
      }
      
      # add modified counts for cases
      data_matrix[sel, cases] = vals_to_mod
      
      # add modified assay back to se
      SummarizedExperiment::assay(se) <- data_matrix
      
      fit_zib_updated = strainspy:::update_fit(fit = fit_zib_null, se = se, update_idx = sel, nthreads = 10)
      fit_ob_updated = strainspy:::update_fit(fit = fit_ob_null, se = se, update_idx = sel, nthreads = 10)
      fit_cascon_updated = strainspy:::update_fit(fit = fit_cascon_null, se = se, update_idx = sel, nthreads = 10, min_identity = 0.9)
      cat("eval_time (min) =", Sys.time() - t0, "sf =", scale_factor, "zp =",  zero_prop, "\n")
      
      op_full = rbind(op_full, 
                      cbind(zero_prop, scale_factor,
                            rbind(
                              c("zib", strainspy:::get_confusion_mx(top_hits = top_hits(fit_zib_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                              c("ob", strainspy:::get_confusion_mx(top_hits = top_hits(fit_ob_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)),
                              c("cc", strainspy:::get_confusion_mx(top_hits = top_hits(fit_cascon_updated), gt_contigs = contigs, all_contigs = rownames(se), print_cm = T)))))
    }
  }
  colnames(op_full) = c("z_prop", "sf", "Test", "tp", "fn", "fp", "tn")
  saveRDS(op_full, save_path)
}
gg_heatmaps(op_full)
```
# Generate a manhattan plot with ground truth for a single data point
```{r}
scale_factor = 0.9
zero_prop = 0.6

exp_beta = c()
exp_zi = c()
se = data # NULL

data_matrix <- SummarizedExperiment::assay(se) # NULL
vals_to_mod = data_matrix[sel, cases]

for(i in 1:length(sel)){ 
  tmp = rescale_beta(x = data_matrix[sel[i], cases]/100,
                     beta = scale_factor,
                     zi = zero_prop)
  
  vals_to_mod[i,] = tmp$rescaled*100
  exp_beta[i] = tmp$expected_beta
  exp_zi[i] = tmp$expected_zi
}

# add modified counts for cases
data_matrix[sel, cases] = vals_to_mod

# add modified assay back to se
SummarizedExperiment::assay(se) <- data_matrix

fit_zib_updated = strainspy:::update_fit(fit = fit_zib_null, se = se, update_idx = sel, nthreads = 10)
fit_ob_updated = strainspy:::update_fit(fit = fit_ob_null, se = se, update_idx = sel, nthreads = 10)
fit_cascon_updated = strainspy:::update_fit(fit = fit_cascon_null, se = se, update_idx = sel, nthreads = 10, min_identity = 0.9)

strainspy:::plot_manhattan_gt(fit_zib_updated, method = "holm", ground_truth = contigs)
strainspy:::plot_manhattan_gt(fit_ob_updated, method = "holm", ground_truth = contigs)
strainspy:::plot_manhattan_gt(fit_cascon_updated, method = "holm", ground_truth = contigs)

```

