---
title: "varyingPrior"
output: github_document
editor_options: 
  chunk_output_type: console
---

# Effects of varying the MAP prior

# Analysing a large-ish dataset (157 non-spiked vs. 20 spiked)

## Load data and prepare for simulations

```{r}
library(strainspy)
library(dplyr)
# Profile 95 - fastest to run
d_p95 <- read_sylph("ecoli_sims/test/sylph_p_95.tsv")

# taxonomy
tax_95 = read_taxonomy(system.file("extdata", "example_taxonomy.tsv.gz", package = "strainspy"))
design <- as.formula(" ~ spiked")

```

### Full model with default weak prior normal(0,5)
```{r, fig.width=12}
# spiked samples
ss = readLines("ecoli_sims/test/sampled.txt")

# metadata 
meta_data = data.frame(run_accession = colnames(d_p95),spiked = as.factor(sapply(colnames(d_p95), function(x) x%in% ss)))

d_p95_full = modify_metadata(se = d_p95, meta_data = meta_data)

d_p95_full <- filter_by_presence(d_p95_full, min_nonzero = 20)

save_path = "output_rds/prior_zib95_weak.rds"
if(file.exists(save_path)){
  full_zib_95_w <- readRDS(save_path)
} else {
  system.time({
    full_zib_95_w <- glmZiBFit(d_p95_full, design, nthreads = 10)
  })
  saveRDS(full_zib_95_w, save_path)
}

top_hits(full_zib_95_w)
# true/false positive
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_w, alpha = 0.05)$Contig_name, drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_w, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')
```

Here, *Clostridiales bacterium* is a false positive. Based on the box plot, it looks like an edge case with just 3 samples driving the signal.

### Full model with strong prior normal(0,1)
```{r, fig.width=12}
save_path = "output_rds/prior_zib95_strong.rds"
if(file.exists(save_path)){
  full_zib_95_s <- readRDS(save_path)
} else {
  system.time({
    full_zib_95_s <- glmZiBFit(d_p95_full, design, nthreads = 10, MAP_prior = "preset_strong")
  })
  saveRDS(full_zib_95_s, save_path)
}


top_hits(full_zib_95_s)

# true/false positive
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_s, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')
```

This fixes the FP issue.

### Full model with ebayes prior
```{r, fig.width=12}

save_path = "output_rds/prior_zib95_ebp.rds"

ebp_full = compute_eb_priors(d_p95_full, design = design, nthreads = parallel::detectCores(), low_cutoff = 0, high_cutoff = Inf)
ebp_full
plot_prior_bootstrap(object = ebp_full, "spikedTRUE")

if(file.exists(save_path)){
  full_zib_95_eb <- readRDS(save_path)
} else {
  system.time({
    full_zib_95_eb <- glmZiBFit(d_p95_full, design, nthreads = 10, MAP_prior = ebp_full)
  })
  saveRDS(full_zib_95_eb, save_path)
}

top_hits(full_zib_95_eb)
# This plot is redundant - same as (0,1) prior
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_s, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')

```

Still no FP as expected. However, eBayes priors were unnecessary for this case. Larger datasets tend to work okay with either weak or strong prior options. Faster to use and maybe more stable.

# Small sample test

This is where ebp priors shine! Here we sample and take 16 spiked vs. 17 non spiked.

## Subset data and prepare for simulations
```{r}
sample_sz = 33

# hack to convert to function
n1 <- floor(sample_sz * 0.5) # Let's do a half split
n2 <- sample_sz - n1

set.seed(1988)
meta_subset <- bind_rows(
  meta_data %>% filter(spiked == TRUE) %>% slice_sample(n = n1),
  meta_data %>% filter(spiked == FALSE) %>% slice_sample(n = n2)
)

d_subset = d_p95[, meta_subset$run_accession]
d_subset <- filter_by_presence(d_subset, min_nonzero = ceiling(max(sample_sz*0.1, 2)))

d_subset = modify_metadata(d_subset, meta_subset)
dim(d_subset)
```

## fit ZIB weak prior

```{r, fig.width=12}
save_path = "output_rds/prior_zib_small_weak.rds"

if(file.exists(save_path)){
  fit_zib_w <- readRDS(save_path)
} else {
  system.time({
    fit_zib_w <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores())
  })
  saveRDS(fit_zib_w, save_path)
}

head(top_hits(fit_zib_w))
plot_manhattan(fit_zib_w, taxonomy = tax_95, aggregate_by_taxa = F)

# Plenty of edge cases
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_w, alpha = 0.05)$Contig_name[1:15], drop_zeros = T, show_points = T, plot_type = 'box')
```

## fit ZIB strong prior

```{r, fig.width=12}
save_path = "output_rds/prior_zib_small_strong.rds"

if(file.exists(save_path)){
  fit_zib_s <- readRDS(save_path)
} else {
  system.time({
    fit_zib_s <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores(), MAP_prior = "preset_strong")
  })
  saveRDS(fit_zib_s, save_path)
}

head(top_hits(fit_zib_s))
plot_manhattan(fit_zib_s, taxonomy = tax_95, aggregate_by_taxa = F)

# Still edge cases
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_s, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')
```

### compare the two priors

```{r}
# At a strong multiple testing correction, both priors pick the same edge cases, but in a different order
all(top_hits(fit_zib_s, alpha = 0.05)$Contig_name %in% top_hits(fit_zib_w, alpha = 0.05)$Contig_name)
plot(1:15, match(top_hits(fit_zib_s, alpha = 0.05)$Contig_name, top_hits(fit_zib_w, alpha = 0.05)$Contig_name))

# But if we relax the multple correction method to BH, the weak prior has more false positives than the strong prior
length(top_hits(fit_zib_w, alpha = 0.05, method = 'BH')$Contig_name)
length(top_hits(fit_zib_s, alpha = 0.05, method = 'BH')$Contig_name) 
# this is as expected

```

Both methods are not great for the small sample size.

## fit empirical bayes prior
```{r, fig.width=12}
ebp = compute_eb_priors(d_subset, design = design, nthreads = parallel::detectCores(), low_cutoff = 0, high_cutoff = Inf)
ebp
plot_prior_bootstrap(ebp, "spikedTRUE")

save_path = "output_rds/prior_zib_small_ebp.rds"

if(file.exists(save_path)){
  fit_zib_eb <- readRDS(save_path)
} else {
  system.time({
    fit_zib_eb <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores(), MAP_prior = ebp)
  })
  saveRDS(fit_zib_eb, save_path)
}


head(top_hits(fit_zib_eb))
plot_manhattan(fit_zib_eb, taxonomy = tax_95, aggregate_by_taxa = F)

# We only have the ZI signal for E. coli now! 

# valid case
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_eb, alpha = 0.05)$Contig_name, drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_eb, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')

```

Using the eBayes prior fixed the false positives