---
title: "varyingPrior"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Effects of varying the MAP prior

# Analysing a large-ish dataset (157 non-spiked vs. 20 spiked)

## Load data and prepare for simulations

```{r}
library(strainspy)
library(dplyr)
# Profile 95 - fastest to run
d_p95 <- read_sylph("ecoli_sims/test/sylph_p_95.tsv")

# taxonomy
tax_95 = read_taxonomy(system.file("extdata", "example_taxonomy.tsv.gz", package = "strainspy"))
design <- as.formula(" ~ spiked")

```

### Full model with default weak prior normal(0,5)
```{r}
# spiked samples
ss = readLines("ecoli_sims/test/sampled.txt")

# metadata 
meta_data = data.frame(run_accession = colnames(d_p95),spiked = as.factor(sapply(colnames(d_p95), function(x) x%in% ss)))

d_p95_full = modify_metadata(se = d_p95, meta_data = meta_data)

d_p95_full <- filter_by_presence(d_p95_full, min_nonzero = 20)
full_zib_95_w <- glmZiBFit(d_p95_full, design, nthreads = 10)
top_hits(full_zib_95_w)
# true/false positive
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_w, alpha = 0.05)$Contig_name, drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_p95_full, phenotype = 'spiked', contigs = top_hits(full_zib_95_w, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')
```

Here, *Clostridiales bacterium* is a false positive. Based on the box plot, it looks like an edge case with just 3 samples driving the signal.

### Full model with strong prior normal(0,1)
```{r}
full_zib_95_s <- glmZiBFit(d_p95_full, design, nthreads = 10, MAP_prior = "preset_strong")
top_hits(full_zib_95_s)
```

This fixes the FP issue.

### Full model with ebayes prior
```{r}
ebp_full = compute_eb_priors(d_p95_full, design = design, nthreads = parallel::detectCores(), low_cutoff = 0, high_cutoff = Inf)
plot_prior_bootstrap(object = ebp_full, "spikedTRUE")

full_zib_95_eb <- glmZiBFit(d_p95_full, design, nthreads = 10, MAP_prior = ebp_full)
top_hits(full_zib_95_eb)
```

Still no FP, but eBayes priors were unnecessary. 

Larger datasets tend to work okay with either weak or strong prior options. Faster to use and maybe more stable.

## Small sample test

This is where our priors shine! Here we sample and take 16 spiked vs. 17 non spiked. The zero-inflated component does not have power to detect signal in smaller samples.

## Subset data and prepare for simulations
```{r}
sample_sz = 33

# hack to convert to function
n1 <- floor(sample_sz * 0.5) # Let's do a half split
n2 <- sample_sz - n1

set.seed(1988)
meta_subset <- bind_rows(
  meta_data %>% filter(spiked == TRUE) %>% slice_sample(n = n1),
  meta_data %>% filter(spiked == FALSE) %>% slice_sample(n = n2)
)

d_subset = d_p95[, meta_subset$run_accession]
d_subset <- filter_by_presence(d_subset, min_nonzero = ceiling(max(sample_sz*0.1, 2)))

d_subset = modify_metadata(d_subset, meta_subset)
```

# fit ZIB weak prior

```{r}
fit_zib_w <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores())
head(top_hits(fit_zib_w))
plot_manhattan(fit_zib_w, taxonomy = tax_95, aggregate_by_taxa = F)
# edge cases
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_w, alpha = 0.05)$Contig_name[1:5], drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_w, alpha = 0.05)$Contig_name[1:15], drop_zeros = T, show_points = T, plot_type = 'box')
```

# fit ZIB strong prior
```{r}
fit_zib_s <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores(), MAP_prior = "preset_strong")
head(top_hits(fit_zib_s))
plot_manhattan(fit_zib_s, taxonomy = tax_95, aggregate_by_taxa = F)
# edge cases
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_s, alpha = 0.05)$Contig_name[1:5], drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_s, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')
```

# fit empirical prior
```{r}
ebp = compute_eb_priors(d_subset, design = design, nthreads = parallel::detectCores(), low_cutoff = 0, high_cutoff = Inf)
plot_prior_bootstrap(ebp, "spikedTRUE")

fit_zib_eb <- glmZiBFit(d_subset,  design, nthreads=parallel::detectCores(), MAP_prior = ebp)

head(top_hits(fit_zib_eb))
plot_manhattan(fit_zib_eb, taxonomy = tax_95, aggregate_by_taxa = F)

# valid case
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_eb, alpha = 0.05)$Contig_name, drop_zeros = F, show_points = T, plot_type = 'box')
plot_ani_dist(d_subset, phenotype = 'spiked', contigs = top_hits(fit_zib_eb, alpha = 0.05)$Contig_name, drop_zeros = T, show_points = T, plot_type = 'box')

```
